### 2018-03-30 paper/paper.Rmd
boilerplate

---
title: S-Tipping vs B-Tipping
## We can figure out author order later on who end up doing more etc :-)
author:
  - name: Carl Boettiger
    email: cboettig@berkeley.edu
    affiliation: espm
    footnote: Corresponding Author
  - name: Ryan Batt
    affiliation: epa
address:
  - code: Some Institute of Technology
    address: Department of Environmental Science, Policy, and Management, University of California, Berkeley, CA 94720
  - code: epa
    address: Environmental Protection Agency, ...
abstract: |
  This is the abstract.

  It consists of two paragraphs.

journal: Journal of Mathematical Biology
bibliography: mybibfile.bib
output: rticles::elsevier_article

keywords: tipping point, bifurcation, saddle-node, state
---

### 2018-03-30 paper/paper.Rmd
sketching outline and more boilerplate

@@
 
-  It consists of two paragraphs.
-
+layout:  3p # review = doublespace, 3p = singlespace, 5p = two-column
+preamble: |
+  \newcommand{\ud}{\mathrm{d}}
+  \usepackage[nomarkers]{endfloat}
+  \linenumbers
+  \usepackage{setspace}
+  \doublespacing
 journal: Journal of Mathematical Biology
-bibliography: mybibfile.bib
+bibliography: refs.bib
+csl: journal-of-mathematical-biology.csl
 output: rticles::elsevier_article
@@
 ---
+
+# Introduction
+
+Critical transistions.
+
+Two flavors: S tipping vs B tipping
+
+Dimensionality.
+Absense of an explicit model often makes it difficult to intuit if and how a complex ecosystem can be intepreted in terms of this simple one-dimensional bifurcation process.
+Lack of clarity about what defines the "state" and what defines the "parameters" leads to confusion.
+These issues are easily glossed over...
+
+Focus on a prominent example from lake ecosystems.
+
+# Model
+
+@Carpenter2008
+
+# Analysis
+
+Dimensionality reduction
+
+## Implications for Resilience and Early Warning
+
+# Discussion
+
+# Acknowledgements
+
+CB acknowledges support in part from USDA National Institute of Food and Agriculture, Hatch project CA-B-INS-0162-H.
+
+\newpage
+
+# References

### 2018-03-30 paper/paper.Rmd
sort

@@
 author:
+  - name: Ryan Batt
+    affiliation: epa
   - name: Carl Boettiger
@@
     footnote: Corresponding Author
-  - name: Ryan Batt
-    affiliation: epa
+
 address:

### 2018-04-20 paper/paper.Rmd
started jotting down eqns

@@
 address:
-  - code: Some Institute of Technology
+  - code: espm
     address: Department of Environmental Science, Policy, and Management, University of California, Berkeley, CA 94720
@@
 
-Dimensionality reduction
+Variable | Description
+-------- | ----------------
+ $A$     | Adult Bass
+ $J$     | Juvenile Bass
+ $F$     | Planktivore zooplankton
+ $H$     | Herbivorous zooplankton
+ $P$     | Phytoplankton
+
+The original model presented in @Carpenter2008 is defined in terms of continuous dynamics within years coupled to discrete annual dynamics.  Within a given year:
+
+$$\frac{d A}{dt} = - q E A$$
+$$d F = \left(D_f (F_R - F) - c_{FA} FA \right) dt  + \sigma_F dW_f$$
+
+$$ \frac{dJ}{dt} = -c_{JA} JA - \frac{c_{JF} v FJ}{h + v + c_{JF} F}$$
+
+While dynamics between years are defined in terms of integrals over the within-year dynamics:
+
+$$A_{t+1} = s \int_t^{t+i} \left(A_{t} + J_{t} \right) dt $$
+$$F_{t+1} = \int_t^{t+i} F_{t} dt $$
+
+$$J_{t+1} = fsA_{t+1}$$
+
+Based on these equations, we can approximate these dynamics in continuous time as:
+
+$$\frac{d A}{dt} = - q E A_t + sJ $$
+$$d F = \left(D_f (F_R - F) - c_{FA} FA \right) dt  + \sigma_F dW_f$$
+
+$$\frac{dJ}{dt} = -c_{JA} JA - \frac{c_{JF} v FJ}{h + v + c_{JF} F} + fA$$
+
+Dynamics for herbivorous plankton and phytoplankton are given only in continuous time:
+
+$$dH = \left(D_H (H_R - H) + \alpha c_{HP} HP - c_{HF} HF\right) dt + \sigma_H dW_h$$
+
+$$dP = \left( r_P L \gamma(I_o, P) P  - m P - c_{PH} HP \right) dt + \sigma_P dW_P$$
+
+### Non-Dimensionalization
+
+$$\tilde F := D_f (F - F_R) $$
+$$\frac{d A}{dt} = - q E A_t + sJ $$
+$$d \tilde F = \left(-\tilde F - c_{fA} \tilde F A \right) dt  + \sigma_F dW_f$$
+$$\alpha := \frac{h+v}{c_{JF}} $$
+
+$$\frac{d J}{dt} = fA -\left(c_{JA} A + \frac{v F}{\alpha + F} \right) J $$
+
+$$\tilde H = D_H (H - H_R) $$
+$$d \tilde H = \tilde H \left(1 + \alpha c_{\tilde HP} P - c_{\tilde HF} F\right) dt + \sigma_H dW_h$$
+
+### Dimensionality reduction
 

### 2018-04-21 paper/paper.Rmd
work in progress

@@
 $$\frac{d A}{dt} = - q E A_t + sJ $$
-$$d \tilde F = \left(-\tilde F - c_{fA} \tilde F A \right) dt  + \sigma_F dW_f$$
-$$\alpha := \frac{h+v}{c_{JF}} $$
+$$d \tilde F = - \left(1 + c_{fA} A \right) \tilde F  dt  + \sigma_F dW_f$$
+$$\beta := \frac{h+v}{c_{JF}} $$
 
-$$\frac{d J}{dt} = fA -\left(c_{JA} A + \frac{v F}{\alpha + F} \right) J $$
+$$\frac{d J}{dt} = fA -\left(c_{JA} A + \frac{v F}{\beta + F} \right) J $$
 
-$$\tilde H = D_H (H - H_R) $$
+$$\tilde H := D_H (H - H_R) $$
 $$d \tilde H = \tilde H \left(1 + \alpha c_{\tilde HP} P - c_{\tilde HF} F\right) dt + \sigma_H dW_h$$

### 2018-05-16 paper/paper.Rmd
i cant use epa affiliation unless my advisors are on the project .... easier to just do this :)

@@
   - name: Ryan Batt
-    affiliation: epa
+    affiliation: Rensselaer Polytechnic Institute and Rutgers University
   - name: Carl Boettiger

### 2018-05-17 pkgBuild/dependencies.R
add a convenience function for updating dependencies
i do this in my projects, might be a hinderance

@@ -1,0 +1,13 @@
+
+update_dependencies <- function(){
+	devtools::use_package("rmarkdown", type="Imports")
+	devtools::use_package("testthat", type="Imports")
+	devtools::use_package("lintr", type="Imports")
+	devtools::use_package("rticles", type="Imports")
+	devtools::use_package("tidyverse", type="Suggests")
+	
+	devtools::use_package("stats", type="Imports")
+	devtools::use_package("methods", type="Imports")
+	devtools::use_package("rootSolve", type="Imports")
+	devtools::use_package("deSolve", type="Imports")
+}

### 2018-05-18 paper/paper.Rmd

--- 2018-04-21 paper/paper.Rmd
@@
   - name: Ryan Batt
-    affiliation: epa
+    affiliation: rpirutgers
   - name: Carl Boettiger
@@
     address: Department of Environmental Science, Policy, and Management, University of California, Berkeley, CA 94720
-  - code: epa
-    address: Environmental Protection Agency, ...
+  - code: rpirutgers
+    address: Rensselaer Polytechnic Institute and Rutgers University
 abstract: |

### 2018-07-06 R/ecoStep.R
a parameterization from the 2013 oikos paper
should work better than the 2008 eco lett parameterization, because the oikos paper did everything in continuous time

@@ -87,8 +87,8 @@
 # J2biom <- 0.05  # Convert J to kg / ha
 # F2biom <- 1  # Convert F to kg / ha
 
-ecoStep <- function(X, pars=c(qE=1), fA=2, cJA=1E-3, cJF=0.5, cFA=0.3, vuln=1, hide=8, surv=0.5, Fo=100, DF=0.1, Zmix=4, nZ=10, rP=3, Load=0.6, mP=0.1, Ho=1, DH=0.5, cHF=0.1, alf=0.3, cPH=0.25){
-	# sigma=0.1,  A2biom=0.2, J2biom=0.05, F2biom=1
+# ecoStep <- function(X, pars=c(qE=1), fA=2, cJA=1E-3, cJF=0.5, cFA=0.3, vuln=1, hide=8, surv=0.5, Fo=100, DF=0.1, Zmix=4, nZ=10, rP=3, Load=0.6, mP=0.1, Ho=1, DH=0.5, cHF=0.1, alf=0.3, cPH=0.25){
+ecoStep <- function(X, pars=c(qE=0.01), fA=2, cJA=0.1, cJF=0.5, cFA=0.3, vuln=80, hide=80, surv=0.6, Fo=200, DF=0.09, Zmix=4, nZ=10, rP=3, Load=0.6, mP=0.1, Ho=1, DH=0.5, cHF=0.1, alf=0.3, cPH=0.25){
 	with(as.list(c(X,pars)),{
 		# Plankton Dynamics
 		GAMMA <- function(z,Pbar) {

### 2018-07-06 R/stability.R
equations of motion w/ less algebra, more program

@@ -262,4 +262,87 @@ stabClass <- function(qE, pars, func=fs2D, Jvals=seq(0,2E3,length.out=2), Fvals=
 }
 
 
+#' Functions of Motion
+#' 
+#' Provides the motion of each state variable (change in that variable as a function of itself and parameters)
+#' 
+#' @param state0 numeric, value of a state variable (can be scalar or vector, not named)
+#' @param stateVar character, name of state variable whose motion you want
+#' @param pars named vector of parameters, needs to at least include qE
+#' 
+#' @note These functions should essentially replace \code{\link{dFJ_dt_1state}}. The A motion function was not an option in \code{\link{dFJ_dt_1state}}, though. Both A motion and J motion are done without much algebra or substitution; however, getting the equation/ function for F motion requires doing some alebraic substitution to get the A state variable to cancel out. Dividing by A and letting it cancel probably involves some extra assumption that might make the F motion equation/ function invalid in certain areas of state-space.
+#' 
+#' @export
+stateMotion <- function(state0, stateVar=c("A0","F0","J0"), pars){
+	stateVar <- match.arg(stateVar)
+	parsF <- unlist(formals(ecoStep)[c("fA", "cJA", "cJF", "cFA", "vuln", "hide", "surv", "Fo", "DF")])
+	if(missing(pars)){ # if a function requires qE, pars needs to be supplied (e.g., pars=c(qE=0.1))
+		pars <- parsF
+	}else{
+		pars <- c(pars, parsF[!names(parsF)%in%names(pars)])
+	}
+	
+	out <- with(as.list(pars), {
+		# Functions that calculate the state variable value
+		Afun <- function(Jhat){ # gives A as a function of J and parameters
+			surv*Jhat/(qE+1-surv)
+		}
+		Afun2 <- function(Fhat){ # gives A as a function of F and parameters, required some algebra first, though
+			surv*fA/(cJA*(qE+1-s)) - (cJF*vuln*Fhat)/(cJA*(hide+vuln+cJF*Fhat))
+		}
+		Ffun <- function(Ahat){ # gives F as a function of A and parameters
[…]

### 2018-07-06 R/ecoStep.R
update default parameter values

@@ -34,12 +34,17 @@
 #' @param cHF consumption rate by planktivore (of zooplankton)
 #' @param alf conversion efficiency of consumed phytoplankton to zooplankton
 #' @param cPH consumption rate of phytoplankton by zooplankton
+#' @param I0 surface irradiance
+#' @param k_sat per microEinsteins m-1 s-1
+#' @param k_inh per microEinsteins
+#' @param DOC dissolved organic carbon, g m-3
+#' @param eps0,epsP extinction coefficient, per m
 #' 
 #' @note \code{F} are referred to as planktivorous fish, but they do feed on the juvenile piscivores (\code{J}).
 #' 
-#' @return named vector of length 3, with names corresponding to juvenile bass (J0), adult bass (A0), and sunfish (F0) abundances.
+#' @return named vector of length 5, with names corresponding to juvenile bass (J0), adult bass (A0), and sunfish (F0) abundances.
 #' @examples
-#' ecoStep(X=c(A0=1.5,F0=25,J0=2), pars=c(qE=0.1))
+#' ecoStep(X=c(A0=1.5,F0=25,J0=2, H0=1, P0=1), pars=c(qE=0.01))
 #' @export
 
 # Phytoplankton Parameters
@@ -88,14 +93,14 @@
 # F2biom <- 1  # Convert F to kg / ha
 
 # ecoStep <- function(X, pars=c(qE=1), fA=2, cJA=1E-3, cJF=0.5, cFA=0.3, vuln=1, hide=8, surv=0.5, Fo=100, DF=0.1, Zmix=4, nZ=10, rP=3, Load=0.6, mP=0.1, Ho=1, DH=0.5, cHF=0.1, alf=0.3, cPH=0.25){
-ecoStep <- function(X, pars=c(qE=0.01), fA=2, cJA=0.1, cJF=0.5, cFA=0.3, vuln=80, hide=80, surv=0.6, Fo=200, DF=0.09, Zmix=4, nZ=10, rP=3, Load=0.6, mP=0.1, Ho=1, DH=0.5, cHF=0.1, alf=0.3, cPH=0.25){
[…]

### 2018-07-06 R/fishStep.R
tell fishStep to get its defaults from ecoStep

@@ -11,16 +11,16 @@
 #' Compute the rate of change of the three fish state variables given initial conditions and parameters
 #' 
 #' @param X vector of length 3 with names A0, F0, and J0.  Indicates starting value of state variables
-#' @param pars a length 1 vector with name qE, indicating harvesting rate of adults
-#' @param fA  Fecundity of adult piscivore (2 in OLD)
-#' @param cJA Density dependent mortality rate of juveniles
-#' @param cJF Consumption of juveniles by planktivores
-#' @param cFA Consumption of planktivores by adult piscivores
-#' @param vuln Vulnerability coefficient (this is "v" in eco lett table/ equations)
-#' @param hide Hiding coefficient (this is "h" in eco lett table/ equations)
-#' @param surv Overwinter survivorship of adults
-#' @param Fo Refuge density of planktivores 
-#' @param DF Diffusion parameter for planktivores
+#' @param pars a named vector with at least qE, indicating harvesting rate of adults; can include other relevant parameters, see \code{\link{ecoStep}}
+# #' @param fA  Fecundity of adult piscivore (2 in OLD)
+# #' @param cJA Density dependent mortality rate of juveniles
+# #' @param cJF Consumption of juveniles by planktivores
+# #' @param cFA Consumption of planktivores by adult piscivores
+# #' @param vuln Vulnerability coefficient (this is "v" in eco lett table/ equations)
+# #' @param hide Hiding coefficient (this is "h" in eco lett table/ equations)
+# #' @param surv Overwinter survivorship of adults
+# #' @param Fo Refuge density of planktivores
+# #' @param DF Diffusion parameter for planktivores
 #' 
 #' @return named vector of length 3, with names corresponding to juvenile bass (J0), adult bass (A0), and sunfish (F0) abundances.
 #' @examples
@@ -72,7 +72,14 @@
 # J2biom <- 0.05  # Convert J to kg / ha
 # F2biom <- 1  # Convert F to kg / ha
 
+# fishStep <- function(X, pars=c(qE=1), fA=2, cJA=1E-3, cJF=0.5, cFA=0.3, vuln=1, hide=8, surv=0.5, Fo=100, DF=0.1){
[…]

### 2018-07-06 R/getRoot.R
add 5D root finder, update examples for new params
ugh, all the examples are ~broken b/c they used to try to show alt stable states etc, but now the values are totally thrown off

@@ -8,10 +8,9 @@
 #' @param ... additional arguments to pass to \code{\link{fishStep}}
 #' @return numeric vector with roots
 #' @examples
-#' getRoot(c(A0=1, F0=1, J0=1), pars=c(qE=0.65))
-#' getRoot(c(A0=1000, F0=1, J0=1000), pars=c(qE=0.65))
-#' getRoot(c(A0=1, F0=1, J0=1), pars=c(qE=0.5))
-#' getRoot(c(A0=1000, F0=1, J0=1000), pars=c(qE=0.5))
+#' getRoot(c(A0=15, F0=50, J0=10), pars=c(qE=-0.02)) # stocking, high starting A
+#' getRoot(c(A0=2, F0=50, J0=5), pars=c(qE=-0.02)) # stocking, but start low A
+#' getRoot(c(A0=15, F0=1, J0=15), pars=c(qE=0.02)) # harvesting, start high A
 #' @export
 getRoot <- function(x, pars, maxiter=1E3, ...){
 	# make function check x and pars for qE
@@ -32,6 +31,39 @@ getRoot <- function(x, pars, maxiter=1E3, ...){
 	}, warning=function(w)c(A0=NA,F0=NA,J0=NA))
 }
 
+#' Get the Root of the Full Ecosystem Model in 5D
+#' 
+#' Numerically finds the roots of the ecosystem model given starting values (adults, [sun]fish, juveniles, herbivores, and phytoplankton) and parameter (harvest of adults, qE)
+#' 
+#' @param x a vector of length 5 with names A0 (adult bass), F0 (planktivorous fish),  J0 (juvenile bass), H0 (herbivorous zooplankton), and P0 (phytoplankton)
+#' @param pars vector of parameters to pass to model
+#' @param maxiter maximum number of iterations to use when finding root
+#' @param ... additional arguments to pass to \code{\link{ecoStep}}
+#' @return numeric vector with roots
+#' @examples
+#' getRoot5D(c(A0=15, F0=50, J0=10, H0=100, P0=100), pars=c(qE=-0.02)) # stocking, high starting A
+#' getRoot5D(c(A0=2, F0=50, J0=5, H0=100, P0=100), pars=c(qE=-0.02)) # stocking, but start low A
+#' getRoot5D(c(A0=15, F0=1, J0=15, H0=100, P0=100), pars=c(qE=0.02)) # harvesting, start high A
+#' @export
[…]

### 2018-07-06 R/ecoStep.R
fiddle with parameter definitions and adjust notation

@@ -79,14 +79,14 @@
 
 # Fish Parameters
 # fA <- 2  # Fecundity of adult piscivore (2 in OLD)
-# cJA <- 1E-3 #0.1  # Density dependent mortality rate of juveniles
+# cJA <- 0.1 #1E-3   # Density dependent mortality rate of juveniles
 # cJF <- 0.5  # Consumption of juveniles by planktivores
 # cFA <- 0.3  # Consumption of planktivores by adult piscivores
-# vuln <- 1 #80  # Vulnerability coefficient (this is "v" in eco lett table/ equations)
-# hide <- 8 #80  # Hiding coefficient (this is "h" in eco lett table/ equations)
-# surv <- 0.5 #0.6  # Overwinter survivorship of adults
-# Fo <- 100  # Refuge density of planktivores  # 100 in OLD
-# DF <- 0.1 #0.09  # Diffusion parameter for planktivores
+# vuln <- 80 #1  # Vulnerability coefficient (this is "v" in eco lett table/ equations)
+# hide <- 80 #8  # Hiding coefficient (this is "h" in eco lett table/ equations)
+# surv <- 0.6 #0.5 #0.6  # Overwinter survivorship of adults
+# Fo <- 200 #100  # Refuge density of planktivores  # 100 in OLD
+# DF <- 0.09 # 0.1  # Diffusion parameter for planktivores
 # sigma <- 0.1 #0.05  # SD of additive noise for planktivores (0.1 in May 07)
 # A2biom <- 0.2  # Convert A to kg / ha
 # J2biom <- 0.05  # Convert J to kg / ha
@@ -107,7 +107,8 @@ ecoStep <- function(X, pars=c(qE=0.01), fA=2, cJA=0.1, cJF=0.5, cFA=0.3, vuln=80
 		
 		# Fish dynamics
 		#Arate <- (surv/nint)*J0 - qE*A0 - ((1-surv)/nint)*A0
-		Arate <- (surv)*J0 - qE*A0 - ((1-surv))*A0
+		# Arate <- (surv)*J0 - qE*A0 - ((1-surv))*A0
+		Arate <- (surv)*J0 - qE*A0 - (mA<-0.4)*A0
 		Frate <- DF*(Fo-F0) - cFA*F0*A0
 		Jpredloss <- (-cJA*J0*A0)-(cJF*vuln*J0*F0/(hide + vuln + cJF*F0) ) # Note this is negative
 	

### 2018-07-06 R/fishStep.R
update fishStep to use ecoStep defaults (correctly this time)

@@ -73,7 +73,7 @@
 # F2biom <- 1  # Convert F to kg / ha
 
 # fishStep <- function(X, pars=c(qE=1), fA=2, cJA=1E-3, cJF=0.5, cFA=0.3, vuln=1, hide=8, surv=0.5, Fo=100, DF=0.1){
-fishStep <- function(X, pars=c(qE=1), fA=2, cJA=1E-3, cJF=0.5, cFA=0.3, vuln=1, hide=8, surv=0.5, Fo=100, DF=0.1){
+fishStep <- function(X, pars=c(qE=1)){
 	parsF <- unlist(formals(ecoStep)[c("fA", "cJA", "cJF", "cFA", "vuln", "hide", "surv", "Fo", "DF")])
 	if(missing(pars)){ # if a function requires qE, pars needs to be supplied (e.g., pars=c(qE=0.1))
 		pars <- parsF
@@ -84,7 +84,8 @@ fishStep <- function(X, pars=c(qE=1), fA=2, cJA=1E-3, cJF=0.5, cFA=0.3, vuln=1, 
 	with(as.list(c(X,pars)),{
 		# Fish dynamics
 		#Arate <- (surv/nint)*J0 - qE*A0 - ((1-surv)/nint)*A0
-		Arate <- (surv)*J0 - qE*A0 - ((1-surv))*A0
+		# Arate <- (surv)*J0 - qE*A0 - ((1-surv))*A0
+		Arate <- (surv)*J0 - qE*A0 - 0.6*A0
 		Frate <- DF*(Fo-F0) - cFA*F0*A0
 		Jpredloss <- (-cJA*J0*A0)-(cJF*vuln*J0*F0/(hide + vuln + cJF*F0) ) # Note this is negative
 	

### 2018-07-06 R/stability.R
add a function for equations of motion for all states

@@ -262,13 +262,15 @@ stabClass <- function(qE, pars, func=fs2D, Jvals=seq(0,2E3,length.out=2), Fvals=
 }
 
 
-#' Functions of Motion
+#' Equations of Motion
 #' 
 #' Provides the motion of each state variable (change in that variable as a function of itself and parameters)
 #' 
 #' @param state0 numeric, value of a state variable (can be scalar or vector, not named)
 #' @param stateVar character, name of state variable whose motion you want
 #' @param pars named vector of parameters, needs to at least include qE
+#' 
+#' @return numeric scalar or vector indicating the rate of change of the state variable
 #' 
 #' @note These functions should essentially replace \code{\link{dFJ_dt_1state}}. The A motion function was not an option in \code{\link{dFJ_dt_1state}}, though. Both A motion and J motion are done without much algebra or substitution; however, getting the equation/ function for F motion requires doing some alebraic substitution to get the A state variable to cancel out. Dividing by A and letting it cancel probably involves some extra assumption that might make the F motion equation/ function invalid in certain areas of state-space.
 #' 
@@ -288,7 +290,7 @@ stateMotion <- function(state0, stateVar=c("A0","F0","J0"), pars){
 			surv*Jhat/(qE+1-surv)
 		}
 		Afun2 <- function(Fhat){ # gives A as a function of F and parameters, required some algebra first, though
-			surv*fA/(cJA*(qE+1-s)) - (cJF*vuln*Fhat)/(cJA*(hide+vuln+cJF*Fhat))
+			surv*fA/(cJA*(qE+1-surv)) - (cJF*vuln*Fhat)/(cJA*(hide+vuln+cJF*Fhat))
 		}
 		Ffun <- function(Ahat){ # gives F as a function of A and parameters
 			DF*Fo/(cFA*Ahat+DF)
@@ -322,27 +324,5 @@ stateMotion <- function(state0, stateVar=c("A0","F0","J0"), pars){
 	})
 	return(unname(out))
 }
-# stateMotion(state0=800, pars=c(qE=1.0))
-# stateMotion(state0=1:3, pars=c(qE=1.0))
-# curve(stateMotion(state0=x, pars=c(qE=1.0)), from=0, to=50)
-#
[…]

### 2018-07-07 notebook/1d-tipping.Rmd

---
title: "Tipping"
output: pdf_document
---

```{r include=FALSE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE, echo=FALSE, fig.height = 3)
```

```{r}
library(tidyverse)
# install_github("thomasp85/patchwork")
library(patchwork)
```

# May Model

May's classic bistable model can be written as follows:

$$\frac{dX_{t}}{dt} =  \underbrace{X_t r \left(1 -\frac{X_t}{K} \right)}_{\textrm{Vegitation growth}}  - \underbrace{\frac{a X_t ^ Q}{X_t^ Q + H ^ Q}}_{\textrm{Vegitation consumption}} + \xi_t,$$

We sometimes call the $dX/dt$ equation the "equation of motion".  Note that it can be thought of as the difference between two separate curves, growth and consumption.  It's often easier to see how we can create a bifurcation when we plot these two curves separately.  Here we consider the plot of the separate curves, the overall equation of motion, and the so-called "Potential Well" or ball-in-cup diagram, all as a function of the state, $X_t$.  Recall that if the equation of motion is defined as $\frac{dX_{t}}{dt} = f(X_t)$ then the "potential" is defined as: $U(X) = - \int_0^X f(Y) dY$.

Let's compute some numbers with some example parameters:

```{r}
p <- list(r = .5, K = 2, Q = 5, H = .38, sigma = .04, a = 0.245)

theory <- 
  tibble(x= seq(0,2, length.out = 100)) %>%
  # Growth and consumption parts separately
  mutate(growth = x * p$r * (1 - x / p$K), 
         consumption = p$a * x ^ p$Q / (x^p$Q + p$H^p$Q)) %>%
  ## 
  mutate(x_dot = growth - consumption) %>%
  ## "Potential" is the negative integral
  mutate(potential = - cumsum(growth - consumption))

```

```{r }
p1 <- theory %>% select(x, growth, consumption) %>%
  gather(curve, y, -x) %>%
  ggplot(aes(x, y, col=curve)) +
  geom_line(lwd=1) +
  labs(y = bquote(dX[t]/dt), x = bquote(X[t])) +
  theme(legend.position = c(0.8, 0.2))
```

```{r }
p2 <- theory %>% select(x, y = x_dot) %>%
  ggplot(aes(x, y)) +
  geom_line(lwd=1) + 
  coord_cartesian(xlim=c(-.1, 1.5)) + 
  geom_hline(yintercept=0, lty=3) + 
  labs(y = bquote(dX[t]/dt), x = bquote(X[t]))
```

```{r}
p3 <- theory %>%
  ggplot(aes(x, potential)) + 
  geom_line(lwd=1) + 
  coord_cartesian(xlim=c(0,1.6), ylim=c(-1.15,-0.9))
```

```{r}
p1 + p2 + p3
```

# Tipping:

Note it is easy to observe how increasing the parameter `a` slightly causes this to "tip":

```{r}
p <- list(r = .5, K = 2, Q = 5, H = .38, sigma = .04, a = 0.255)

theory <- 
  tibble(x= seq(0,2, length.out = 100)) %>%
  # Growth and consumption parts separately
  mutate(growth = x * p$r * (1 - x / p$K), 
         consumption = p$a * x ^ p$Q / (x^p$Q + p$H^p$Q)) %>%
  ## 
  mutate(x_dot = growth - consumption) %>%
  ## "Potential" is the negative integral
  mutate(potential = - cumsum(growth - consumption))

```

```{r }
p1 <- theory %>% select(x, growth, consumption) %>%
  gather(curve, y, -x) %>%
  ggplot(aes(x, y, col=curve)) +
  geom_line(lwd=1) +
  labs(y = bquote(dX[t]/dt), x = bquote(X[t]))+
  theme(legend.position = c(0.8, 0.2))
```

```{r }
p2 <- theory %>% select(x, y = x_dot) %>%
  ggplot(aes(x, y)) +
  geom_line(lwd=1) + 
  coord_cartesian(xlim=c(-.1, 1.5)) + 
  geom_hline(yintercept=0, lty=3) + 
  labs(y = bquote(dX[t]/dt), x = bquote(X[t]))
```

```{r}
p3 <- theory %>%
  ggplot(aes(x, potential)) + 
  geom_line(lwd=1) + labs(subtitle = "Potential Well")
```

```{r}
p1 + p2 + p3
```

### 2018-07-12 pkgBuild/manuscript_report.R
update report to include composite eqs of motion AFJ

@@ -122,10 +122,10 @@ opts_chunk$set(
 #+ setup
 set.seed(42)
 qE <- -0.001 #0.02
-qE_end <- 0.001 #-5#-0.025
+qE_end <- 0.001#/20 #-5#-0.025
 
 dt <- 0.01 # 0.01
-nYears <- 1020
+nYears <- 1020#/20
 
 noise_coeff <- c(0.01, 0.01, 0.01, 0.01, 0.01)
 
@@ -202,6 +202,7 @@ for(i in 1:length(qevals)){
 }
 #' As with the adults, the second stable node appears when qE>0
 #'   
+#'   
 #' \FloatBarrier  
 #'   
 #' ***  
@@ -215,13 +216,26 @@ for(i in 1:length(qevals)){
 	abline(h=0, lty=2)
 	mtext(paste0("qE=",round(qevals[i],2)), side=3, line=0, adj=0.05, font=2)
 }
-#' Unfortuneately, this state variable only shows a trivial stable node and a saddle point. This is the state variable that requires algebra cancelling the 'A' state variable in order to get its equation of motion (without A canceling, you cannot get a dF/dt vs F type plot easily [as a function of F and parameters]). We think this result is invalid, as there must be some assumptions in the algebra that invalidate our desired interpretation.  
+#' Unfortunately, this state variable only shows a trivial stable node and a saddle point. This is the state variable that requires algebra cancelling the 'A' state variable in order to get its equation of motion (without A canceling, you cannot get a dF/dt vs F type plot easily [as a function of F and parameters]). We think this result is invalid, as there must be some assumptions in the algebra that invalidate our desired interpretation.  
 #'   
 #' \FloatBarrier  
 #'   
 #' ***  
 #'   
-
+#'   
+#' ##Figure: Combined A, J, F Motion
[…]

### 2018-07-13 pkgBuild/manuscript_report.R
use m_A instead of 1-s; use c_{} in more places

@@ -90,27 +90,27 @@ opts_chunk$set(
 #' 
 #' Focusing on the fish, and setting each of the 3 fish differential equations to 0 and solving for the respective state variables:  
 #' \begin{array}{llr}
-#' A&= \frac{sJ}{qE+1-s} &(6)\\[10pt]
+#' A&= \frac{sJ}{qE+m_A} &(6)\\[10pt]
 #' J&=\frac{fA}{c_{JA}A+s+\frac{c_{JF}vF}{h+v+c_{JF}F}} &(7)\\[10pt]
 #' F&=\frac{D_FF_0}{D_F + c_{FA}A} &(8)\\[10pt]
 #' \end{array}
 #' 
 #' Substitute Eq7 into Eq6, and solve for $A$:  
 #' \begin{array}{llr}
-#' A&=\frac{sf}{c_{JA}(qE+1-s)} - \frac{s}{c_{JA}} - \frac{c_{JF}vF}{x(h+v+c_{JF}F)} &(9)
+#' A&=\frac{sf}{c_{JA}(qE+m_A)} - \frac{s}{c_{JA}} - \frac{c_{JF}vF}{c_{JA}(h+v+c_{JF}F)} &(9)
 #' \end{array}
 #' 
 #' Substitute Eq9 into Eq2, giving us the dynamics of $F$ as a function of $F$ and parameters:  
 #' \begin{array}{llr}
-#' \dot{F} &= d(F_o-F) - c_{FA}F(\frac{sf}{c_{JA}(qE+1-s)} - \frac{s}{c_{JA}} - \frac{c_{JF}vF}{c_{JA}(h+v+c_{JF}F)}) &(10)
+#' \dot{F} &= D_F(F_o-F) - c_{FA}F(\frac{sf}{c_{JA}(qE+m_A)} - \frac{s}{c_{JA}} - \frac{c_{JF}vF}{c_{JA}(h+v+c_{JF}F)}) &(10)
 #' \end{array}
 #'   
 #' Eq10 is what is used in `dF_dt_1state`.  
 #' 
 #' **Alternatively**, we can rearrange the equations to perform a different seat of substituions to solve for $\dot{J}$ as a function of $J$ and parameters (juvenile bass):  
 #' \begin{array}{llr}
-#' Q &\equiv 1/(qE + 1 - s) \\
[…]

### 2018-07-13 pkgBuild/manuscript_report.R
more intermediates for eqs of motion, clean notation

@@ -91,26 +91,36 @@ opts_chunk$set(
 #' Focusing on the fish, and setting each of the 3 fish differential equations to 0 and solving for the respective state variables:  
 #' \begin{array}{llr}
 #' A&= \frac{sJ}{qE+m_A} &(6)\\[10pt]
-#' J&=\frac{fA}{c_{JA}A+s+\frac{c_{JF}vF}{h+v+c_{JF}F}} &(7)\\[10pt]
-#' F&=\frac{D_FF_0}{D_F + c_{FA}A} &(8)\\[10pt]
+#' F&=\frac{D_FF_0}{D_F + c_{FA}A} &(7)\\[10pt]
+#' J&=\frac{fA}{c_{JA}A+s+\frac{c_{JF}vF}{h+v+c_{JF}F}} &(8)\\[10pt]
 #' \end{array}
 #' 
-#' Substitute Eq7 into Eq6, and solve for $A$:  
+
+#' 
+#' **Begin Equations of Motion**  
+#' Although our ultimate goal is to see if the equation of motion for $J$ or $F$ shows alternate states, for completeness we will also present the equation of motion for $A$, which is acquired by substituting the RHS of Eq7 for $F$ in Eq8, then substituting the resulting expression for $J$ in Eq1:  
 #' \begin{array}{llr}
-#' A&=\frac{sf}{c_{JA}(qE+m_A)} - \frac{s}{c_{JA}} - \frac{c_{JF}vF}{c_{JA}(h+v+c_{JF}F)} &(9)
+#' \dot{A} &=  sfA\Bigg(c_{JA}+s+c_{JF}v\bigg(\frac{(h+v)(D_F+c_{FA}A)}{D_F F_o}+c_{JF}\bigg)^{-1}\Bigg)^{-1} &(9)
 #' \end{array}
 #' 
-#' Substitute Eq9 into Eq2, giving us the dynamics of $F$ as a function of $F$ and parameters:  
+#' To get the equation of motion for $F$, we begin by getting an expression for $A$ that depends only on $F$ and parameters, which is achieved by substituting Eq8 into Eq6 and solving for $A$:
 #' \begin{array}{llr}
[…]

### 2018-07-13 paper/paper.Rmd
Rmd version

--- 2018-04-21 paper/paper.Rmd
@@
 ---
-title: S-Tipping vs B-Tipping
-## We can figure out author order later on who end up doing more etc :-)
+title: "Bifurcation or state tipping: assessing transition type in a model trophic cascade"
 author:
   - name: Ryan Batt
-    affiliation: epa
+    affiliation: rb
   - name: Carl Boettiger
@@
     footnote: Corresponding Author
-
 address:
   - code: espm
-    address: Department of Environmental Science, Policy, and Management, University of California, Berkeley, CA 94720
-  - code: epa
-    address: Environmental Protection Agency, ...
+    address: "Department of Environmental Science, Policy, and Management, University of California, Berkeley, CA 94720"
+  - code: rb
+    address: "Rensselaer Polytechnic Institute, Department of Biological Sciences, Troy, NY 12309 USA; 2Rutgers University, Department of Ecology, Evolution, and Natural Resources, New Brunswick, NJ 08901 USA"
+
 abstract: |
@@
 
-Critical transistions.
+Ecological systems prone to abrupt changes, such as a fishery collapsing, can be challenging to manage.
+These abrupt dynamics are often referred to as regime shifts, and result from nonlinearities in the underlying system dynamics.
+Therefore, ecologists have often turned to studying the dynamics of nonlinear but simple mathematical models to better understand and manage ecosystems.
 
-Two flavors: S tipping vs B tipping
+There are essentially two different mechanisms by which a system exhibiting alternative stable states can experience a sudden transition from one state to another [@Scheffer2001]: state tipping (S-tipping) or bifurcation (B-tipping).
+Figure 1 illustrates these two different mechanisms.
+B-tipping is a sudden transition that is induced by slow changes to a parameter of the underlying dynamical system that reduce the stability and ultimately destroy the stable node, as it collides with an unstable saddle-point in a saddle-node bifurcation (aka a fold bifurcation or "blue sky catastrophe," [@Strogatz2001]).
+In contrast, S-tipping involves a perturbation directly to a state variable that carries the system across the separatrix into an alternative stable state, without any such change to the underlying shape or existence of either attractor.
+S-tipping can be driven by purely stochastic events or direct exogenous forcing.
+Crucially, only one of these mechanisms, B-tipping, will exhibit the phenomenon of Critical Slowing Down (CSD), a pattern that has received great attention in both theoretical and empirical studies for its potential as an early warning signal of such a regime shift [e.g.
+@Wissel1984; @Scheffer2009; @Scheffer2015].
 
-Dimensionality.
-Absense of an explicit model often makes it difficult to intuit if and how a complex ecosystem can be intepreted in terms of this simple one-dimensional bifurcation process.
-Lack of clarity about what defines the "state" and what defines the "parameters" leads to confusion.
-These issues are easily glossed over...
+Despite this attention, convincing empirical demonstrations of critical slowing down are still rare, particularly outside carefully controlled laboratory settings [e.g.
+@Dai2012].
+A foundational example can be found in studies of lake ecosystems [e.g.
+@Carpenter2008], which can exist in either a "piscivore dominated" or "planktivore dominated" states (Figure 1).
+Significant previous work has established evidence for these alternative stable states and the potential for sudden transitions between the states that cannot be easily reversed.
+Then in 2011, a landmark study of @Carpenter2011 analyzed the experimental manipulation of an initially plantivore-dominated state as a definitive example of CSD phenomena providing an early warning signal prior to a sudden transition to the piscivore dominated state.
+Yet was this manipulation truly an example of B-tipping, where CSD is theoretically predicted, or S-tipping, which does not create the CSD phenomenon?
 
-Focus on a prominent example from lake ecosystems.
+B-tipping is typically considered to be the result of slow change in a parameter of [e.g.
+@Scheffer2009] the system which ultimately leads to the fold bifurcation (or close enough to a fold bifurcation for stochasticity to finish the job).
+S-tipping involves manipulation of the state variable.
+@Carpenter2011 describe driving the critical transition through the gradual addition of piscivores (largemouth Bass), which drives a transition characterized by low-piscivore density into one of high piscivore density.
+If the state (piscivore density) is being manipulated directly, then is this not an example of S-tipping, and thus unsuitable to test for the presence of CSD?
+Although the simplicity of ball-in-cup (potential well) based descriptions of regime shifts makes them alluring, this question is impossible to address properly without a more mathematically precise description of the system and the manipulation involved.
+Surprisingly, despite the central importance the piscivore transitions have played in our understanding of ecosystem regime shifts, these details have never been clearly presented in prior publications.
+Here, we provide a direct analysis of the underlying model in the piscivory alternative stable state system to demonstrate why these studies are indeed an example of B-tipping.
 
-# Model
+Central to resolving this question is untangling a second issue that frequently arises in the discussion of alternative stable state dynamics in ecological systems: the issue of dimension.
+The S-tipping and B-tipping transitions illustrated in Figure 1 are fundamentally rooted in one-dimensional analysis: the state consists of a single continuously valued dynamical variable (i.e.
+population density of adult bass).
+Ecosystems are notoriously more complex, and this lake ecosystem is no exception.
+Here, we use analyze previously published models  of trophic cascades to determine whether manipulating the top predator constitutes B-tipping or S-tipping.
+Common versions include 5 state variables, ranging from top predators to primary producers.
+We accomplish this by performing dimension reduction on the full 5D model, and surmising whether the reduced model can be expressed in a way such that manipulating predator abundance is tantamount to manipulating a parameter, or if it necessarily constitutes state manipulation.
+This exercise facilitates the interpretation of the lake experiment in the context of predicting regime shifts by using a 1D model to understand a high dimensional system.
 
-@Carpenter2008
+# Methods
 
-# Analysis
+## Model Description
 
-Variable | Description
--------- | ----------------
- $A$     | Adult Bass
- $J$     | Juvenile Bass
- $F$     | Planktivore zooplankton
- $H$     | Herbivorous zooplankton
- $P$     | Phytoplankton
+In the model of the lake food web, largemouth bass (Micropterus salmoides, A) are the top predator.
+Intuitively, juvenile bass (J) mature into A, and J are produced by A.
+Bass feed on other fish, including their own young (J) and other planktivorous fish (F), (i.e.
+Sunfish and similar forage fish species).
+These planktivores and juvenile bass (which are also planktivores) eat herbivorous zooplankton (H), which in turn eat phytoplankton (P).
+Because bass fry are small enough to be eaten by "planktivorous" fish, J can be consumed by F.
+Furthermore, F and J compete with each other for food (H).
+Thus, there are strong pairwise interactions between the three fish groups, and the dynamics of these groups extend into the planktonic portion of the food web.
+	The five-dimensional model of the food web includes the aforementioned state variables consisting of three fish groups (A, F, J) and two planktonic groups (H, P):
 
-The original model presented in @Carpenter2008 is defined in terms of continuous dynamics within years coupled to discrete annual dynamics.  Within a given year:
+$$\begin{array}{llr} \dot{A} &= sJ - qA - m_A A &(1) \\ \dot{F} &= D_F(F_o - F) - c_{FA}FA &(2)\\ \dot{J} &= fA - c_{JA}JA - \frac{c_{JF}vJF}{h+v+c_{JF}F} - sJ &(3)\\ \dot{H} &= D_H(H_R-H) + \alpha c_{HP}HP - c_{HF}HF &(4)\\ \dot{P} &= r_P L \gamma (I_0,P)P - mP - c_{PH}HP &(5)\\ \end{array}$$
 
-$$\frac{d A}{dt} = - q E A$$
-$$d F = \left(D_f (F_R - F) - c_{FA} FA \right) dt  + \sigma_F dW_f$$
+Decreases in adult bass abundance result from mortality ($m_A A$) and harvest ($qA$), and increases result from the maturation of the fraction of $J$ that survive ($s$) mortality (Eq.
+1).
+Planktivorous fish diffuse ($D_F$) between a refuge of size $F_o$ and a foraging arena, and the predation rate of $A$ on $F$ is controlled by $c_{FA}$ (Eq.
+2).
+Growth in $J$ is controlled by the fecundity ($f$) of $A$, but are subject to losses via cannibalization at a rate $c_{JA}$, and interactions with $F$ which are controlled by the rate at which $J$ hide ($h$) in a refuge or become vulnerable ($v$) to predation at a rate $c_{JF}$ as they enter an implicit foraging arena (Eq.
+3).
+Herbivorous zooplankton diffuse between surface waters and a deep water refuge at a rate $D_H$, grow by preying on phytoplankton at rate cHP and converting phytoplankton biomass (phosphorus) by a factor $\alpha$, and decline as they are preyed upon by planktivores at a rate $c_{HF}$ (Eq.
+4).
+Phytoplankton death is due to mortality and sinking at rate m or predation by herbivores at rate cHF, while their growth is controlled by the per capita growth rate ($r_P$), the availability of phosphorus (introduced at rate $L$), and the availability of light at the surface ($I_0$) and the density dependent growth response to phytoplankton in response to light ($\gamma$, Eq.
+5).
 
-$$ \frac{dJ}{dt} = -c_{JA} JA - \frac{c_{JF} v FJ}{h + v + c_{JF} F}$$
+## Dimension Reduction
 
-While dynamics between years are defined in terms of integrals over the within-year dynamics:
+Our approach to determining whether the lake experiment constituted B-tipping or S-tipping is to reduce the dimensionality of the model described in Eqs.
+1-5, and see if a 1-D expression of this model can produce alternate states and the corresponding fold bifurcation.
+Although transitions occur throughout the food web, we focus here on the fish dynamics (Eqs.
+1-3) as they are most directly related to either harvest (q) or direct manipulation of adult bass (A).
+In the model, the dynamics of herbivorous zooplankton (H) and phytoplankton (P) do not influence A, F, or J.
+Therefore, the dimensionality of the model can easily be reduced from 5 to 3 by ignoring Eqs.
+4-5 (the plankton) without loss of generality for the dynamics of Eqs.
+1-3 (the fish).
+In addition to parameters, the dynamics of each fish state depend on itself and at least one other fish state.
+We observe that the dynamics of both the Juvenile bass (J) and forage fish (F) occur on fast timescales relative to the dynamics of long-lived adult bass.
+Therefore, we set Eqs.
+2-3 to zero and solved for the estimate of the respective state variables given equilibrium:
 
-$$A_{t+1} = s \int_t^{t+i} \left(A_{t} + J_{t} \right) dt $$
-$$F_{t+1} = \int_t^{t+i} F_{t} dt $$
+$$\begin{array}{llr}
+\hat A&= \frac{sJ}{q+m_A} &(6)\\
+\hat F&=\frac{D_FF_0}{D_F + c_{FA}A} &(7)\\
+\hat J&=\frac{fA}{c_{JA}A+s+\frac{c_{JF}vF}{h+v+c_{JF}F}} &(8)\\
+\end{array}$$
 
-$$J_{t+1} = fsA_{t+1}$$
+The equation of motion for A (an expression of A's dynamics completely in terms of A and fixed parameters), is acquired by substituting the RHS of Eq.
+7 for F in Eq.
+8, then substituting the resulting expression for J in Eq.
+1.
+After some substitution, we can write the equation of motion for the adult bass, A, completely in terms of the adult bass population and fixed parameters of the system.
 
-Based on these equations, we can approximate these dynamics in continuous time as:
+$$\begin{array}{llr}
+\dot{A} &=  sfA\Bigg(c_{JA}A+s+c_{JF}v\bigg(\frac{(h+v)(D_F+c_{FA}A)}{D_F F_o}+c_{JF}\bigg)^{-1}\Bigg)^{-1} - qA - m_A A &(9)
+\end{array}$$
 
-$$\frac{d A}{dt} = - q E A_t + sJ $$
-$$d F = \left(D_f (F_R - F) - c_{FA} FA \right) dt  + \sigma_F dW_f$$
+To determine if this equation exhibits alternative stable states and the potential for B-tipping with changing $q$
 
-$$\frac{dJ}{dt} = -c_{JA} JA - \frac{c_{JF} v FJ}{h + v + c_{JF} F} + fA$$
+## Simulations
 
-Dynamics for herbivorous plankton and phytoplankton are given only in continuous time:
+We performed simulations in which we initialized the 5D model at equilibrium with a low but negative harvest rate (q=-0.001), which can be interpreted as per capita stocking of A.
+We proceeded to linearly increase q at each time step over 1020 years with a time step size of 0.01 years (102000 time steps total) until q reached a value of 0.001.
+At each time step a small amount of additive, normally distributed noise was added to Eqs.
+1-3.
+Values for state variables less than 0.001 were set to 0.001 to prevent ecologically impossible values resulting from our numerical approximation of continuous dynamics.
+R code for these simulations is included in supplementary materials.
 
-$$dH = \left(D_H (H_R - H) + \alpha c_{HP} HP - c_{HF} HF\right) dt + \sigma_H dW_h$$
+# Results
 
-$$dP = \left( r_P L \gamma(I_o, P) P  - m P - c_{PH} HP \right) dt + \sigma_P dW_P$$
+Plotting this equation using established ranges for ecological parameters [CITATION],  Figure 3 shows that the system should support two alternative stable states, and can undergo a fold bifurcation (in reverse, creating a new stable point) when "catch" q is decreased sufficiently.
+This sudden transition is felt in other dimensions as it's impact cascades down to the lower trophic levels.
 
-### Non-Dimensionalization
+These same dynamics can be seen in the direct numerical simulations of the full 5-dimensional model.
+At a q of 0, the fish in the simulation underwent a critical transition (Figure 2).
+This transition occurred during time step 510, although an abrupt change in state was not observed until step 850.
+The changes in A and J were similar, but were inversely related to those in F.
+The predation effect of A on J was small compared to the role of A's fecundity in producing J and A's role as a predator of F, which negatively impacts J as well.
+Overall, both bass time series decreased with increasing q (abruptly when stocking converted to harvest at q=0), whereas the abundance of planktivorous fish was positively correlated with $q$.
 
-$$\tilde F := D_f (F - F_R) $$
-$$\frac{d A}{dt} = - q E A_t + sJ $$
-$$d \tilde F = - \left(1 + c_{fA} A \right) \tilde F  dt  + \sigma_F dW_f$$
-$$\beta := \frac{h+v}{c_{JF}} $$
-
-$$\frac{d J}{dt} = fA -\left(c_{JA} A + \frac{v F}{\beta + F} \right) J $$
-
-$$\tilde H := D_H (H - H_R) $$
-$$d \tilde H = \tilde H \left(1 + \alpha c_{\tilde HP} P - c_{\tilde HF} F\right) dt + \sigma_H dW_h$$
-
-### Dimensionality reduction
-
-## Implications for Resilience and Early Warning
+The fish dynamics cascaded down to lower trophic levels, although H and P did not undergo abrupt transitions at the same time as the fishes.
+Instead, $H$ and $P$ exhibited strong nonlinear changes much earlier in the time series after $q$ had increased only slightly.
+The small cumulative increase in $q$ leading up to step 20 caused a slight increase in $F$ and a dramatic decline in $H$ ($F$'s prey); after its initial decline, the abundance of $H$ was 0.001 for the remainder of the time series.
+When H abruptly declined, P abruptly increased, then continued to increased approximately linearly for the remainder of the time series.
+Therefore, increasing q caused declines in both bass states ($A$ and $J$), an increase in planktivorous fish ($F$), a decline in herbivorous zooplankton ($H$), and an increase in phytoplankton ($P$), a pattern consistent with a trophic cascade.
+All state variables exhibited abrupt state changes at some point in the time series.
 
 # Discussion
+
+These results demonstrate that the system considered can indeed be driven through fold bifurcations as the parameter $q$ is slowly tuned up or down, forcing the creation or annihilation of an alternative stable state -- confirming the presence of B-tipping in this system.
+Simulations of the same system using previous estimates for ecological parameters confirm that the predictions of the 1-dimensional approximation (eq 9) correspond with the dynamics observed in the full model.
+
+This analysis also highlights the importance of what might appear a merely semantic issue that in fact corresponds to fundamentally different dynamics.
+In this formulation, the control parameter q corresponds to harvest (positive q) or stocking (negative q) of adult bass in the lake.
+It is this aspect that first gives the appearance of state tipping dynamics, since it suggests the state is being somehow "directly manipulated".
+Yet from a mathematical standpoint, it is irrelevant if the control parameter corresponds to some environmental or policy variable (fishing regulations, fish migration into the lake), which in turn increases or decreases net mortality proportionally, or if the manipulation directly adds or removes adult bass -- the feature that distinguishes this manipulation from a S-tipping transition comes down to the timescales involved.
+Slowly increasing in the amount of bass stocked in the lake (i.e.
+decreasing q, effectively running the scenario of increasing fishing harvests in reverse) gives the rest of the food web time to adjust.
+In contrast, a single large introduction of many bass would create an S-tipping transition, driving the system immediately into an alternate stable state without exhibiting CSD phenomena needed to evaluate early warning signals of a tipping point.
+
+All though we have focused on the analysis of deterministic models for simplicity, these results can easily be interpreted in the context of stochastic models as well.
+(Recall that in practice, the detection of early warning signals associated with CSD requires some level of stochasticity to generate the variance needed to detect slowing down).
+Under a stochastic model, the distinction between S-tipping and B-tipping is slightly blurred, since a system that gets close enough to the tipping point due to changing parameters (B-tipping) will ultimately be able to jump across the separatrix (S-tipping) before the actual bifurcation event.
+In a stochastic system, fluctuations to the state variable act as intrinsic sources of perturbations, but though part of the system, large chance deviations are no more predictable than sudden exogenous forcing: there is no CSD prior to purely stochastic transitions [@PRSB2013], even though retrospectively a large deviation will share similar statistical signatures such as rapidly increasing variance and autocorrelation [@PRSB2012].
+
+More generally, our analysis draws attention to issues of dimensionality that have been largely overlooked despite significant interest in questions of regime shifts and early warning signals [e.g.
+@Clements2018; @Scheffer2015; @Scheffer2012].
+The popularity of this topic has sometimes led to casually equating any sudden shift with fold bifurcations, overlooking any assumptions necessary to approximate a complex dynamical system with such a one-dimensional model.
+Influential reviews on abrupt change in ecological systems [e.g.
+@Ratajczak2018] focus almost exlusively on descriptions of one-dimensional state-spaces without discussion of if, when or how complex systems such as the ecosystem food-web considered here can be appropriately described by a non-linear equation in a single dimension.
+The analysis shown here illustrates this process on the basis of a fast-slow timescale approximation, which allows us to replace the faster dynamics at lower trophic levels into their equilibrium states.
+(Note also that the nonlinearity necessary to support alternate stable states only emerges through this approximation.)  Thus, for the five-dimensional model considered here, our analysis shows that the familiar one-dimensional characterization of a sudden shift being driven by a fold bifurcation is indeed appropriate, but this was not all-together evident a priori.
+The fact that CSD phenomena can occur in many different models has been suggested to mean that CSD is a generic or universal early warning indicator [@Scheffer2009] of a critical transition.
+In practice, many higher dimensional models may not be so well approximated along a single manifold, or may correspond to other bifurcations or transistions [@TE2013].
+As such, analyses of regime shifts should not be too quick to sweep these details under the rug.
+Mechanistic understanding will always require models.
 
@@
 CB acknowledges support in part from USDA National Institute of Food and Agriculture, Hatch project CA-B-INS-0162-H.
+The authors also acknowledge Alan Hastings, whose off-hand question at a seminar talk inspired the analysis presented here.
 

### 2018-07-13 paper/paper.Rmd
figure

@@
 
+![S-tipping vs B-tipping.
+A critical transition in a system with alternative stable states can be driven either by a perturbation to the state, driving the system across the separatrix (top panel), or by a bifurcation, in which a stable point is annihilated (bottom panel)](figures/scheffer-fig.png)
+
 B-tipping is typically considered to be the result of slow change in a parameter of [e.g.
@@
 3).
-Herbivorous zooplankton diffuse between surface waters and a deep water refuge at a rate $D_H$, grow by preying on phytoplankton at rate cHP and converting phytoplankton biomass (phosphorus) by a factor $\alpha$, and decline as they are preyed upon by planktivores at a rate $c_{HF}$ (Eq.
+Herbivorous zooplankton diffuse between surface waters and a deep water refuge at a rate $D_H$, grow by preying on phytoplankton at rate $c_{HP}$ and converting phytoplankton biomass (phosphorus) by a factor $\alpha$, and decline as they are preyed upon by planktivores at a rate $c_{HF}$ (Eq.
 4).

### 2018-07-13 pkgBuild/manuscript_report.R
add A motion ball/cup, bifurc diag, and param table
and fix a motion equation

@@ -100,7 +100,7 @@ opts_chunk$set(
 #' **Begin Equations of Motion**  
 #' Although our ultimate goal is to see if the equation of motion for $J$ or $F$ shows alternate states, for completeness we will also present the equation of motion for $A$, which is acquired by substituting the RHS of Eq7 for $F$ in Eq8, then substituting the resulting expression for $J$ in Eq1:  
 #' \begin{array}{llr}
-#' \dot{A} &=  sfA\Bigg(c_{JA}+s+c_{JF}v\bigg(\frac{(h+v)(D_F+c_{FA}A)}{D_F F_o}+c_{JF}\bigg)^{-1}\Bigg)^{-1} &(9)
+#' \dot{A} &=  sfA\Bigg(c_{JA}A+s+c_{JF}v\bigg(\frac{(h+v)(D_F+c_{FA}A)}{D_F F_o}+c_{JF}\bigg)^{-1}\Bigg)^{-1} -qA -m_A A &(9)
 #' \end{array}
 #' 
 #' To get the equation of motion for $F$, we begin by getting an expression for $A$ that depends only on $F$ and parameters, which is achieved by substituting Eq8 into Eq6 and solving for $A$:
@@ -127,6 +127,148 @@ opts_chunk$set(
 #'   
 #' ***  
 #'   
+#' 
+#' #Table of Parameter Values
+#+ parameterTable, results='markup', echo=FALSE
+
+# Symbol
+# Description
+# Quantity
+# Units
+paramDefs <- c(
+# "A",
+# "Adult bass",
+# "variable",
+# "kg ha^{-1}", # so, it's actually number of fish, but if you multiply by 0.2 kg/fish you get kg/ha
+#
+# "F",
+# "Planktivorous fish",
+# "variable",
+# "kg ha^{-1}", # actually n, but divide by 500 to get kg/fish
+#
+# "J",
+# "Juvenile bass",
+# "variable",
+# "kg ha^{-1}", # actually n, but divide by 10 to get kg/ha
+#
+# "H",
+# "Herbivorous zooplankton",
+# "variable",
+# "", # no biomass conversion listed in carpenter et al 2008 table
+#
+# "P",
+# "Phytoplankton",
+# "variable",
[…]

### 2018-07-14 pkgBuild/manuscript_report.R
fix figure dims

@@ -388,8 +388,13 @@ for(s in 1:3){
 		mtext(paste0("qE=",round(qevals[i],2)), side=3, line=0, adj=0.05, font=2)
 	}
 }
+#'   
+#' \FloatBarrier  
+#'   
+#' ***  
+#'   
 #' ##Figure: A Motion A Ball-in-Cup
-#+ figure-Amotion-ballincup, fig.width=6, height=3.5, fig.cap="**Figure.** Motion of A, and its integral (ball-in-cup diagram)."
+#+ figure-Amotion-ballincup, fig.width=6, fig.height=3.5, fig.cap="**Figure.** Motion of A, and its integral (ball-in-cup diagram)."
 Adot <- function(A, pars=c(qE=-0.001)){
 	parsF <- unlist(formals(ecoStep)[c("fA", "cJA", "cJF", "cFA", "vuln", "hide", "surv", "Fo", "DF")])
 	if(missing(pars)){ # if a function requires qE, pars needs to be supplied (e.g., pars=c(qE=0.1))

### 2018-07-14 paper/paper.Rmd
add 4 figures

@@
 
-![S-tipping vs B-tipping.
+![Figure 1: S-tipping vs B-tipping.
 A critical transition in a system with alternative stable states can be driven either by a perturbation to the state, driving the system across the separatrix (top panel), or by a bifurcation, in which a stable point is annihilated (bottom panel)](figures/scheffer-fig.png)
@@
 Here, we use analyze previously published models  of trophic cascades to determine whether manipulating the top predator constitutes B-tipping or S-tipping.
-Common versions include 5 state variables, ranging from top predators to primary producers.
+Common versions include 5 state variables, ranging from top predators to primary producers (Figure 2).
 We accomplish this by performing dimension reduction on the full 5D model, and surmising whether the reduced model can be expressed in a way such that manipulating predator abundance is tantamount to manipulating a parameter, or if it necessarily constitutes state manipulation.
 This exercise facilitates the interpretation of the lake experiment in the context of predicting regime shifts by using a 1D model to understand a high dimensional system.
+
+![Figure 2: Food Web Heuristic.
+Adult bass give rise to juveniles, while eating their own young and pumpkinseed sunfish.
+Sunfish compete with and eat juvenile bass.
+Sunfish eat herbivorous zooplankton like *Daphnia*, which in turn eat phytoplankton.
+](figures/Slide1.png)
 
@@
 
+![Figure 3. Time series of the 5 state variables, with q (harvest rate of adult bass) being increased linearly over time. ](../pkgBuild/manuscript_report/figure-5dsimulation-1.png)
+
 These same dynamics can be seen in the direct numerical simulations of the full 5-dimensional model.
-At a q of 0, the fish in the simulation underwent a critical transition (Figure 2).
+At a q of 0, the fish in the simulation underwent a critical transition (Figure 4).
 This transition occurred during time step 510, although an abrupt change in state was not observed until step 850.
@@
 Overall, both bass time series decreased with increasing q (abruptly when stocking converted to harvest at q=0), whereas the abundance of planktivorous fish was positively correlated with $q$.
+
+![Figure 4.
+Top row: motion of A for different values of q.
+Bottom row: the negative integral of the top row, depecting in the so-called ball-in-cup diagrams.](../pkgBuild/manuscript_report/figure-Amotion-ballincup-1.png)
 
@@
 
-These results demonstrate that the system considered can indeed be driven through fold bifurcations as the parameter $q$ is slowly tuned up or down, forcing the creation or annihilation of an alternative stable state -- confirming the presence of B-tipping in this system.
+These results demonstrate that the system considered can indeed be driven through fold bifurcations as the parameter $q$ is slowly tuned up or down (Figure 5), forcing the creation or annihilation of an alternative stable state -- confirming the presence of B-tipping in this system.
 Simulations of the same system using previous estimates for ecological parameters confirm that the predictions of the 1-dimensional approximation (eq 9) correspond with the dynamics observed in the full model.
+
+![Figure 5.
+Bifurcation diagram derived from eq 9.
+Solid lines are stable equilibria, the dashed line is the unstable equilibrium or separatrix.](../pkgBuild/manuscript_report/figure-bifurcationDiagram-1.png)
 

### 2018-07-14 paper/paper.Rmd
add table, adj cite, cite batt2017 re:s-tipping

@@
 In contrast, S-tipping involves a perturbation directly to a state variable that carries the system across the separatrix into an alternative stable state, without any such change to the underlying shape or existence of either attractor.
-S-tipping can be driven by purely stochastic events or direct exogenous forcing.
+S-tipping can be driven by purely stochastic events or direct exogenous forcing, such as extreme events that are common in the biological variables of ecosystems [@Batt2017].
 Crucially, only one of these mechanisms, B-tipping, will exhibit the phenomenon of Critical Slowing Down (CSD), a pattern that has received great attention in both theoretical and empirical studies for its potential as an early warning signal of such a regime shift [e.g.
@@
 5).
+
+**Table 1.** Model parameters, description, values, and units.
+Parameters are taken from [@carpenter2008] and [@Carpenter2013].
+The function \\gamma describes phytoplankton growth, and can be found in [@carpenter2008].
+
+|Symbol |Description                                                 |Quantity |Units               |
+|:------|:-----------------------------------------------------------|:--------|:-------------------|
+|q      |Harvest or stocking                                         |variable |t^{-1}              |
+|s      |Survival rate of juveniles to maturation                    |0.6      |t^{-1}              |
+|m_A    |Mortality rate of adults                                    |0.4      |t^{-1}              |
+|m      |Phytoplankton mortality                                     |0.1      |t^{-1}              |
+|f      |Fecundity of adult bass                                     |2        |J/A                 |
+|c_{FA} |Predation of planktivores by adult bass                     |0.3      |t^{-1} A^{-1}       |
+|c_{JA} |Predation of juvenile bass on by adult bass                 |0.1      |t^{-1} A^{-1}       |
+|c_{JF} |Predation of juvenile bass by planktivorous fish            |0.5      |t^{-1} F^{-1}       |
+|c_{PH} |Predation of phytoplankton by herbivores                    |0.25     |t^{-1} H^{-1}       |
+|c_{HF} |Predation of herbivores by planktivores                     |0.1      |t^{-1} F^{-1}       |
+|F_o    |Abundance of planktivorous fish in non-foraging arena       |200      |F                   |
+|H_R    |Reservoir of herbivores in a deep-water refuge              |4        |H                   |
+|D_F    |Diffusion of planktivores between refuge and foraging arena |0.09     |t^{-1}              |
+|DH     |Diffusion of between shallow and deep water                 |0.5      |t^{-1}              |
+|v      |Rate at which J enter a foraging arena, becoming vulnerable |80       |t^{-1}              |
+|h      |Rate at which J hide in a refuge                            |80       |t^{-1}              |
+|r_P    |Growth rate of phytoplankton                                |3        |m^2 mg^{-1}         |
+|L      |Phosphorus loading rate                                     |0.6      |mg m^{-2} t^{-1}    |
+|\alpha |Assimilation of phytoplankton phosphorus by zooplankton     |0.3      |g g^{-1}            |
+|\gamma |Density-dependent phytoplankton growth response to light    |n/a      |n/a                 |
+|I0     |Solar irradiance incident on the surface of the water       |300      |\mu E m^{-2} s^{-1} |
 

### 2018-07-14 pkgBuild/manuscript_report.R
add kable caption

@@ -268,7 +268,7 @@ paramDefs <- c(
 )
 colNames <- c('Symbol','Description','Quantity','Units')
 paramTable_mat <- matrix(paramDefs, byrow=TRUE, ncol=4, dimnames=list(NULL,colNames))
-kable(paramTable_mat)
+kable(paramTable_mat, caption="**Table 1.** Model parameters, description, values, and units. Parameters are taken from [@carpenter2008] and [@carpenter2013]. The function \\gamma describes phytoplankton growth, and can be found in [@carpenter2008].")
 
 #' #Setup
 #+ setup

### 2018-07-14 paper/paper.Rmd
add an abstract, tidy up

--- 2018-07-13 paper/paper.Rmd
@@
 abstract: |
-  This is the abstract.
+  Ecosystems can experience sudden regime shifts due to a variety of mechanisms.
+Among these are two different mechanisms for crossing a tipping point: when a perturbation to the system state is large enough to push the system beyond the basin of attraction of one stable state and into that of another (state tipping), and alternately, when slow changes to some underlying parameter lead to a fold bifurcation that anihilates one of the stable states.
+Only the latter process generates the phenomena of critical slowing down (CSD) which have been postulated as potential early warning signs ahead of a sudden shift.
+yet distinguishing between the two mechanisms is not always a straight forward as it might seem.
+The distinction between "state" and "parameter" that is self-evident in mathematical equations depends fundamentally on ecological details in model formulation.
+This is particularly relevant when considering high-dimensional models involving trophic webs of interacting species, which can only be reduced this one-dimensional model of a tipping point under appropriate consideration of mathematics and biology involved.
+Here we illustrate that process for a highly influential trophic cascade model used to demonstrate tipping points and test CSD predictions in a natural ecosystem, resolving a previously unclear issue as to the nature of the tipping point involved.
 
@@
 
-keywords: tipping point, bifurcation, saddle-node, state
+keywords: tipping point, bifurcation, saddle-node, alternative stable states
 ---
@@
 Therefore, ecologists have often turned to studying the dynamics of nonlinear but simple mathematical models to better understand and manage ecosystems.
-
 There are essentially two different mechanisms by which a system exhibiting alternative stable states can experience a sudden transition from one state to another [@Scheffer2001]: state tipping (S-tipping) or bifurcation (B-tipping).
-Figure 1 illustrates these two different mechanisms.
+Figure \ref{fig1} illustrates these two different mechanisms.
 B-tipping is a sudden transition that is induced by slow changes to a parameter of the underlying dynamical system that reduce the stability and ultimately destroy the stable node, as it collides with an unstable saddle-point in a saddle-node bifurcation (aka a fold bifurcation or "blue sky catastrophe," [@Strogatz2001]).
 In contrast, S-tipping involves a perturbation directly to a state variable that carries the system across the separatrix into an alternative stable state, without any such change to the underlying shape or existence of either attractor.
-S-tipping can be driven by purely stochastic events or direct exogenous forcing.
+S-tipping can be driven by purely stochastic events or direct exogenous forcing, such as extreme events that are common in the biological variables of ecosystems [@Batt2017].
 Crucially, only one of these mechanisms, B-tipping, will exhibit the phenomenon of Critical Slowing Down (CSD), a pattern that has received great attention in both theoretical and empirical studies for its potential as an early warning signal of such a regime shift [e.g.
@@
 A foundational example can be found in studies of lake ecosystems [e.g.
-@Carpenter2008], which can exist in either a "piscivore dominated" or "planktivore dominated" states (Figure 1).
+@Carpenter2008], which can exist in either a "piscivore dominated" or "planktivore dominated" states (Figure \ref{fig2}).
 Significant previous work has established evidence for these alternative stable states and the potential for sudden transitions between the states that cannot be easily reversed.
-Then in 2011, a landmark study of @Carpenter2011 analyzed the experimental manipulation of an initially plantivore-dominated state as a definitive example of CSD phenomena providing an early warning signal prior to a sudden transition to the piscivore dominated state.
+Then in 2011, a landmark study of @Carpenter2011 analyzed the experimental manipulation of an initially planktivore-dominated state as a definitive example of CSD phenomena providing an early warning signal prior to a sudden transition to the piscivore dominated state.
 Yet was this manipulation truly an example of B-tipping, where CSD is theoretically predicted, or S-tipping, which does not create the CSD phenomenon?
 
-![S-tipping vs B-tipping.
-A critical transition in a system with alternative stable states can be driven either by a perturbation to the state, driving the system across the separatrix (top panel), or by a bifurcation, in which a stable point is annihilated (bottom panel)](figures/scheffer-fig.png)
+\begin{figure}
+\centering
+\includegraphics[width=.8\linewidth]{figures/scheffer-fig.png}
+\caption{S-tipping vs B-tipping. A critical transition in a system with
+alternative stable states can be driven either by a perturbation to the
+state, driving the system across the separatrix (top panel), or by a
+bifurcation, in which a stable point is annihilated (bottom panel). 
+Figure reproduced from @Scheffer2001 with permission.}
+\end{figure}
 
@@
 Central to resolving this question is untangling a second issue that frequently arises in the discussion of alternative stable state dynamics in ecological systems: the issue of dimension.
-The S-tipping and B-tipping transitions illustrated in Figure 1 are fundamentally rooted in one-dimensional analysis: the state consists of a single continuously valued dynamical variable (i.e.
+The S-tipping and B-tipping transitions illustrated in Figure \ref{fig1} are fundamentally rooted in one-dimensional analysis: the state consists of a single continuously valued dynamical variable (i.e.
 population density of adult bass).
@@
 Here, we use analyze previously published models  of trophic cascades to determine whether manipulating the top predator constitutes B-tipping or S-tipping.
-Common versions include 5 state variables, ranging from top predators to primary producers.
+Common versions include 5 state variables, ranging from top predators to primary producers (Figure \ref{fig2}).
 We accomplish this by performing dimension reduction on the full 5D model, and surmising whether the reduced model can be expressed in a way such that manipulating predator abundance is tantamount to manipulating a parameter, or if it necessarily constitutes state manipulation.
 This exercise facilitates the interpretation of the lake experiment in the context of predicting regime shifts by using a 1D model to understand a high dimensional system.
+
+\begin{figure}
+\centering
+\includegraphics[width=.8\linewidth]{figures/Slide1.png}
+\caption{Food Web Heuristic. Adult bass give rise to juveniles, while
+eating their own young and pumpkinseed sunfish. Sunfish compete with and
+eat juvenile bass. Sunfish eat herbivorous zooplankton like
+\emph{Daphnia}, which in turn eat phytoplankton.}\label{fig2}
+\end{figure}
 
@@
 
-In the model of the lake food web, largemouth bass (Micropterus salmoides, A) are the top predator.
+In the model of the lake food web, largemouth bass (\emph{Micropterus salmoides}, A) are the top predator.
 Intuitively, juvenile bass (J) mature into A, and J are produced by A.
@@
 
-Decreases in adult bass abundance result from mortality ($m_A A$) and harvest ($qA$), and increases result from the maturation of the fraction of $J$ that survive ($s$) mortality (Eq.
+Decreases in adult bass abundance result from mortality ($m_A A$) and harvest ($q$), and increases result from the maturation of the fraction of $J$ that survive ($s$) mortality (Eq.
 1).
@@
 4).
-Phytoplankton death is due to mortality and sinking at rate m or predation by herbivores at rate cHF, while their growth is controlled by the per capita growth rate ($r_P$), the availability of phosphorus (introduced at rate $L$), and the availability of light at the surface ($I_0$) and the density dependent growth response to phytoplankton in response to light ($\gamma$, Eq.
+Phytoplankton death is due to mortality and sinking at rate m or predation by herbivores at rate $c_{HF}$, while their growth is controlled by the per capita growth rate ($r_P$), the availability of phosphorus (introduced at rate $L$), and the availability of light at the surface ($I_0$) and the density dependent growth response to phytoplankton in response to light ($\gamma$, Eq.
 5).
+
+**Table 1.** Model parameters, description, values, and units.
+Parameters are taken from [@Carpenter2008] and [@Carpenter2013].
+The function $\gamma$ describes phytoplankton growth, and can be found in [@Carpenter2008].
+
+|Symbol |Description                                                 |Quantity |Units               |
+|:------|:-----------------------------------------------------------|:--------|:-------------------|
+|$q$      |Harvest or stocking                                         |variable |$t^{-1}$              |
+|$s$      |Survival rate of juveniles to maturation                    |0.6      |$t^{-1}$              |
+|$m_A$    |Mortality rate of adults                                    |0.4      |$t^{-1}$              |
+|$m$      |Phytoplankton mortality                                     |0.1      |$t^{-1}$              |
+|$f$      |Fecundity of adult bass                                     |2        |$J/A$                 |
+|$c_{FA}$ |Predation of planktivores by adult bass                     |0.3      |$t^{-1} A^{-1}$       |
+|$c_{JA}$ |Predation of juvenile bass on by adult bass                 |0.1      |$t^{-1} A^{-1}$       |
+|$c_{JF}$ |Predation of juvenile bass by planktivorous fish            |0.5      |$t^{-1} F^{-1}$       |
+|$c_{PH}$ |Predation of phytoplankton by herbivores                    |0.25     |$t^{-1} H^{-1}$     |
+|$c_{HF}$ |Predation of herbivores by planktivores                     |0.1      |$t^{-1} F^{-1}$       |
+|$F_o$    |Abundance of planktivorous fish in non-foraging arena       |200      |$F$                   |
+|$H_R$    |Reservoir of herbivores in a deep-water refuge              |4        |$H$                   |
+|$D_F$    |Diffusion of planktivores between refuge and foraging arena |0.09     |$t^{-1}$            |
+|$D_H$     |Diffusion of between shallow and deep water                 |0.5      |$t^{-1}$              |
+|$v$      |Rate at which J enter a foraging arena, becoming vulnerable |80       |$t^{-1}$              |
+|$h$      |Rate at which J hide in a refuge                            |80       |$t^{-1}$              |
+|$r_P$    |Growth rate of phytoplankton                                |3        |$m^2 mg^{-1}$         |
+|$L$      |Phosphorus loading rate                                     |0.6      |$mg m^{-2} t^{-1}$    |
+|$\alpha$ |Assimilation of phytoplankton phosphorus by zooplankton     |0.3      |$g g^{-1}$            |
+|$\gamma$ |Density-dependent phytoplankton growth response to light    |n/a      |n/a                 |
+|$I_0$     |Solar irradiance incident on the surface of the water       |300      |$\mu E m^{-2} s^{-1}$ |
 
@@
 Although transitions occur throughout the food web, we focus here on the fish dynamics (Eqs.
-1-3) as they are most directly related to either harvest (q) or direct manipulation of adult bass (A).
+1-3) as they are most directly related to either harvest ($q$) or direct manipulation of adult bass (A).
 In the model, the dynamics of herbivorous zooplankton (H) and phytoplankton (P) do not influence A, F, or J.
@@
 
-To determine if this equation exhibits alternative stable states and the potential for B-tipping with changing $q$
+Solving for the roots of this equation demonstrates alternative stable states and the potential for B-tipping with changing $q$, as illustrated graphically in Figure \ref{fig4}.
 
@@
 
-We performed simulations in which we initialized the 5D model at equilibrium with a low but negative harvest rate (q=-0.001), which can be interpreted as per capita stocking of A.
-We proceeded to linearly increase q at each time step over 1020 years with a time step size of 0.01 years (102000 time steps total) until q reached a value of 0.001.
+We performed simulations in which we initialized the 5D model at equilibrium with a low but negative harvest rate ($q=-0.001$), which can be interpreted as per capita stocking of $A$.
+We proceeded to linearly increase $q$ at each time step over 1020 years with a time step size of 0.01 years (102000 time steps total) until $q$ reached a value of 0.001.
 At each time step a small amount of additive, normally distributed noise was added to Eqs.
@@
 Values for state variables less than 0.001 were set to 0.001 to prevent ecologically impossible values resulting from our numerical approximation of continuous dynamics.
-R code for these simulations is included in supplementary materials.
+R code for these simulations is included in supplementary materials (<https://github.com/cboettig/bs-tipping).
 
@@
 
-Plotting this equation using established ranges for ecological parameters [CITATION],  Figure 3 shows that the system should support two alternative stable states, and can undergo a fold bifurcation (in reverse, creating a new stable point) when "catch" q is decreased sufficiently.
+Plotting this equation using established ranges for ecological parameters (Table 1; @Carpenter2008; @Carpenter2013),  Figure \ref{fig4} shows that the system should support two alternative stable states, and can undergo a fold bifurcation (in reverse, creating a new stable point) when "catch" $q$ is decreased sufficiently.
 This sudden transition is felt in other dimensions as it's impact cascades down to the lower trophic levels.
 
+\begin{figure}
+\centering
+\includegraphics[width=.8\linewidth]{../pkgBuild/manuscript_report/figure-5dsimulation-1.png}
+\caption{Time series of the 5 state variables, with \(q\) (harvest rate
+of adult bass) being increased linearly over time.}\label{fig3}
+\end{figure}
+
 These same dynamics can be seen in the direct numerical simulations of the full 5-dimensional model.
-At a q of 0, the fish in the simulation underwent a critical transition (Figure 2).
+At a $q$ of 0, the fish in the simulation underwent a critical transition (Figure \ref{fig4}).
 This transition occurred during time step 510, although an abrupt change in state was not observed until step 850.
@@
 The predation effect of A on J was small compared to the role of A's fecundity in producing J and A's role as a predator of F, which negatively impacts J as well.
-Overall, both bass time series decreased with increasing q (abruptly when stocking converted to harvest at q=0), whereas the abundance of planktivorous fish was positively correlated with $q$.
+Overall, both bass time series decreased with increasing $q$ (abruptly when stocking converted to harvest at $q = 0$), whereas the abundance of planktivorous fish was positively correlated with $q$.
+
+\begin{figure}
+\includegraphics[width=\linewidth]{../pkgBuild/manuscript_report/figure-Amotion-ballincup-1.png}
+\caption{Top row: motion of A for different values of $q$. Bottom row: the negative integral of the top row, depecting in the so-called ball-in-cup (or potential well) diagrams.} \label{fig4}
+\end{figure}
 
@@
 When H abruptly declined, P abruptly increased, then continued to increased approximately linearly for the remainder of the time series.
-Therefore, increasing q caused declines in both bass states ($A$ and $J$), an increase in planktivorous fish ($F$), a decline in herbivorous zooplankton ($H$), and an increase in phytoplankton ($P$), a pattern consistent with a trophic cascade.
+Therefore, increasing $q$ caused declines in both bass states ($A$ and $J$), an increase in planktivorous fish ($F$), a decline in herbivorous zooplankton ($H$), and an increase in phytoplankton ($P$), a pattern consistent with a trophic cascade.
 All state variables exhibited abrupt state changes at some point in the time series.
@@
 
-These results demonstrate that the system considered can indeed be driven through fold bifurcations as the parameter $q$ is slowly tuned up or down, forcing the creation or annihilation of an alternative stable state -- confirming the presence of B-tipping in this system.
-Simulations of the same system using previous estimates for ecological parameters confirm that the predictions of the 1-dimensional approximation (eq 9) correspond with the dynamics observed in the full model.
+These results demonstrate that the system considered can indeed be driven through fold bifurcations as the parameter $q$ is slowly tuned up or down (Figure \ref{fig5}), forcing the creation or annihilation of an alternative stable state -- confirming the presence of B-tipping in this system.
+Simulations of the same system using previous estimates for ecological parameters confirm that the predictions of the 1-dimensional approximation (Eq 9) correspond with the dynamics observed in the full model.
+
+\begin{figure}
+\centering
+\includegraphics[width=.8\linewidth]{../pkgBuild/manuscript_report/figure-bifurcationDiagram-1.png}
+\caption{Bifurcation diagram derived from Eq 9. Solid lines are stable
+equilibria, the dashed line is the unstable equilibrium or separatrix.}\label{fig5}
+\end{figure}
 
 This analysis also highlights the importance of what might appear a merely semantic issue that in fact corresponds to fundamentally different dynamics.
-In this formulation, the control parameter q corresponds to harvest (positive q) or stocking (negative q) of adult bass in the lake.
+In this formulation, the control parameter $q$ corresponds to harvest (positive $q$) or stocking (negative $q$) of adult bass in the lake.
 It is this aspect that first gives the appearance of state tipping dynamics, since it suggests the state is being somehow "directly manipulated".
@@
 Slowly increasing in the amount of bass stocked in the lake (i.e.
-decreasing q, effectively running the scenario of increasing fishing harvests in reverse) gives the rest of the food web time to adjust.
+decreasing $q$, effectively running the scenario of increasing fishing harvests in reverse) gives the rest of the food web time to adjust.
 In contrast, a single large introduction of many bass would create an S-tipping transition, driving the system immediately into an alternate stable state without exhibiting CSD phenomena needed to evaluate early warning signals of a tipping point.
@@
 Under a stochastic model, the distinction between S-tipping and B-tipping is slightly blurred, since a system that gets close enough to the tipping point due to changing parameters (B-tipping) will ultimately be able to jump across the separatrix (S-tipping) before the actual bifurcation event.
-In a stochastic system, fluctuations to the state variable act as intrinsic sources of perturbations, but though part of the system, large chance deviations are no more predictable than sudden exogenous forcing: there is no CSD prior to purely stochastic transitions [@PRSB2013], even though retrospectively a large deviation will share similar statistical signatures such as rapidly increasing variance and autocorrelation [@PRSB2012].
+In a stochastic system, fluctuations to the state variable act as intrinsic sources of perturbations, but though part of the system, large chance deviations are no more predictable than sudden exogenous forcing: there is no CSD prior to purely stochastic transitions [@PRSB2013], even though retrospectively a large deviation will share similar statistical signatures such as rapidly increasing variance and auto-correlation [@PRSB2012].
 
@@
 Influential reviews on abrupt change in ecological systems [e.g.
-@Ratajczak2018] focus almost exlusively on descriptions of one-dimensional state-spaces without discussion of if, when or how complex systems such as the ecosystem food-web considered here can be appropriately described by a non-linear equation in a single dimension.
+@Ratajczak2018] focus almost exclusively on descriptions of one-dimensional state-spaces without discussion of if, when or how complex systems such as the ecosystem food-web considered here can be appropriately described by a non-linear equation in a single dimension.
 The analysis shown here illustrates this process on the basis of a fast-slow timescale approximation, which allows us to replace the faster dynamics at lower trophic levels into their equilibrium states.
-(Note also that the nonlinearity necessary to support alternate stable states only emerges through this approximation.)  Thus, for the five-dimensional model considered here, our analysis shows that the familiar one-dimensional characterization of a sudden shift being driven by a fold bifurcation is indeed appropriate, but this was not all-together evident a priori.
+(Note also that the non-linearity necessary to support alternate stable states only emerges through this approximation.)  Thus, for the five-dimensional model considered here, our analysis shows that the familiar one-dimensional characterization of a sudden shift being driven by a fold bifurcation is indeed appropriate, but this was not all-together evident a priori.
 The fact that CSD phenomena can occur in many different models has been suggested to mean that CSD is a generic or universal early warning indicator [@Scheffer2009] of a critical transition.
-In practice, many higher dimensional models may not be so well approximated along a single manifold, or may correspond to other bifurcations or transistions [@TE2013].
+In practice, many higher dimensional models may not be so well approximated along a single manifold, or may correspond to other bifurcations or transitions [@TE2013].
 As such, analyses of regime shifts should not be too quick to sweep these details under the rug.
@@
 CB acknowledges support in part from USDA National Institute of Food and Agriculture, Hatch project CA-B-INS-0162-H.
-The authors also acknowledge Alan Hastings, whose off-hand question at a seminar talk inspired the analysis presented here.
+The authors also wish to acknowledge Alan Hastings, whose off-hand question at a seminar talk inspired the question and analysis considered here.
 

### 2018-07-14 paper/paper.Rmd
edit/ proofread manuscript

@@
   Ecosystems can experience sudden regime shifts due to a variety of mechanisms.
-Among these are two different mechanisms for crossing a tipping point: when a perturbation to the system state is large enough to push the system beyond the basin of attraction of one stable state and into that of another (state tipping), and alternately, when slow changes to some underlying parameter lead to a fold bifurcation that anihilates one of the stable states.
-Only the latter process generates the phenomena of critical slowing down (CSD) which have been postulated as potential early warning signs ahead of a sudden shift.
-yet distinguishing between the two mechanisms is not always a straight forward as it might seem.
-The distinction between "state" and "parameter" that is self-evident in mathematical equations depends fundamentally on ecological details in model formulation.
-This is particularly relevant when considering high-dimensional models involving trophic webs of interacting species, which can only be reduced this one-dimensional model of a tipping point under appropriate consideration of mathematics and biology involved.
-Here we illustrate that process for a highly influential trophic cascade model used to demonstrate tipping points and test CSD predictions in a natural ecosystem, resolving a previously unclear issue as to the nature of the tipping point involved.
+Two of the ways a system can cross a tipping point include when a perturbation to the system state is large enough to push the system beyond the basin of attraction of one stable state and into that of another (state tipping), and alternately, when slow changes to some underlying parameter lead to a fold bifurcation that anihilates one of the stable states.
+The first mechanism does not generate the phenomenon of critical slowing down (CSD), whereas the latter does generate CSD, which has been postulated as a way to detect early warning signs ahead of a sudden shift.
+Yet distinguishing between the two mechanisms (s-tipping and b-tipping) is not always as straightforward as it might seem.
+The distinction between "state" and "parameter" that may seem self-evident in mathematical equations depends fundamentally on ecological details in model formulation.
+This distinction is particularly relevant when considering high-dimensional models involving trophic webs of interacting species, which can only be reduced to a one-dimensional model of a tipping point under appropriate consideration of both the mathematics and biology involved.
+Here we illustrate that process of dimension reduction and distinguishing between s- and b-tipping for a highly influential trophic cascade model used to demonstrate tipping points and test CSD predictions in silico, and later, in a natural lake ecosystem.
+Our analysis resolves a previously unclear issue as to the nature of the tipping point involved.
 
@@
 
-keywords: tipping point, bifurcation, saddle-node, alternative stable states
+keywords: tipping point, bifurcation, saddle-node, alternative stable states, trophic cascade, regime shift, trophic triangle
 ---
@@
 Therefore, ecologists have often turned to studying the dynamics of nonlinear but simple mathematical models to better understand and manage ecosystems.
-There are essentially two different mechanisms by which a system exhibiting alternative stable states can experience a sudden transition from one state to another [@Scheffer2001]: state tipping (S-tipping) or bifurcation (B-tipping).
-Figure \ref{fig1} illustrates these two different mechanisms.
+There are essentially two different mechanisms by which a system exhibiting alternative stable states can experience a sudden transition from one state to another [@Scheffer2001]: state tipping (S-tipping) or a bifurcation (B-tipping).
+Figure \ref{fig1} illustrates these two  mechanisms.
 B-tipping is a sudden transition that is induced by slow changes to a parameter of the underlying dynamical system that reduce the stability and ultimately destroy the stable node, as it collides with an unstable saddle-point in a saddle-node bifurcation (aka a fold bifurcation or "blue sky catastrophe," [@Strogatz2001]).
 In contrast, S-tipping involves a perturbation directly to a state variable that carries the system across the separatrix into an alternative stable state, without any such change to the underlying shape or existence of either attractor.
-S-tipping can be driven by purely stochastic events or direct exogenous forcing, such as extreme events that are common in the biological variables of ecosystems [@Batt2017].
+S-tipping can be driven by direct exogenous forcing or purely stochastic events, such as extreme events that are common in the biological time series of ecosystems [@Batt2017].
 Crucially, only one of these mechanisms, B-tipping, will exhibit the phenomenon of Critical Slowing Down (CSD), a pattern that has received great attention in both theoretical and empirical studies for its potential as an early warning signal of such a regime shift [e.g.
@@
 A foundational example can be found in studies of lake ecosystems [e.g.
-@Carpenter2008], which can exist in either a "piscivore dominated" or "planktivore dominated" states (Figure \ref{fig2}).
-Significant previous work has established evidence for these alternative stable states and the potential for sudden transitions between the states that cannot be easily reversed.
+@Carpenter2008], which can exist in either a "piscivore dominated" or "planktivore dominated" state (Figure \ref{fig2}).
+Significant previous work has established evidence for these alternative states and the potential for sudden transitions between the states that cannot be easily reversed.
 Then in 2011, a landmark study of @Carpenter2011 analyzed the experimental manipulation of an initially planktivore-dominated state as a definitive example of CSD phenomena providing an early warning signal prior to a sudden transition to the piscivore dominated state.
@@
 Surprisingly, despite the central importance the piscivore transitions have played in our understanding of ecosystem regime shifts, these details have never been clearly presented in prior publications.
-Here, we provide a direct analysis of the underlying model in the piscivory alternative stable state system to demonstrate why these studies are indeed an example of B-tipping.
+Here, we provide a direct analysis of the underlying model in the piscivory alternative state system to demonstrate why these studies are indeed an example of B-tipping.
 
 Central to resolving this question is untangling a second issue that frequently arises in the discussion of alternative stable state dynamics in ecological systems: the issue of dimension.
-The S-tipping and B-tipping transitions illustrated in Figure \ref{fig1} are fundamentally rooted in one-dimensional analysis: the state consists of a single continuously valued dynamical variable (i.e.
+The S-tipping and B-tipping transitions illustrated in Figure \ref{fig1} are fundamentally rooted in a one-dimensional analysis: the state consists of a single continuously valued dynamical variable (i.e.
 population density of adult bass).
 Ecosystems are notoriously more complex, and this lake ecosystem is no exception.
-Here, we use analyze previously published models  of trophic cascades to determine whether manipulating the top predator constitutes B-tipping or S-tipping.
-Common versions include 5 state variables, ranging from top predators to primary producers (Figure \ref{fig2}).
-We accomplish this by performing dimension reduction on the full 5D model, and surmising whether the reduced model can be expressed in a way such that manipulating predator abundance is tantamount to manipulating a parameter, or if it necessarily constitutes state manipulation.
+Here, we analyze a previously published model of a trophic cascade to determine whether manipulating the top predator constitutes B-tipping or S-tipping.
+Common versions of this model include 5 state variables, ranging from top predators to primary producers (Figure \ref{fig2}).
+We perform dimension reduction on the full 5D model, and surmise whether the reduced model can be expressed in a way such that manipulating predator abundance is tantamount to manipulating a parameter, or if it necessarily constitutes state manipulation.
 This exercise facilitates the interpretation of the lake experiment in the context of predicting regime shifts by using a 1D model to understand a high dimensional system.
@@
 Bass feed on other fish, including their own young (J) and other planktivorous fish (F), (i.e.
-Sunfish and similar forage fish species).
+pumpkinseed sunfish, \emph{Lepomis gibbosus} and similar forage fish species).
 These planktivores and juvenile bass (which are also planktivores) eat herbivorous zooplankton (H), which in turn eat phytoplankton (P).
@@
 3).
-Herbivorous zooplankton diffuse between surface waters and a deep water refuge at a rate $D_H$, grow by preying on phytoplankton at rate $c_{HP}$ and converting phytoplankton biomass (phosphorus) by a factor $\alpha$, and decline as they are preyed upon by planktivores at a rate $c_{HF}$ (Eq.
+Herbivorous zooplankton diffuse between surface waters and a deep water refuge (of size $H_R_$) at a rate $D_H$, grow by preying on phytoplankton at rate $c_{HP}$ and converting phytoplankton biomass (phosphorus) by a factor $\alpha$, and decline as they are preyed upon by planktivores at a rate $c_{HF}$ (Eq.
 4).
-Phytoplankton death is due to mortality and sinking at rate m or predation by herbivores at rate $c_{HF}$, while their growth is controlled by the per capita growth rate ($r_P$), the availability of phosphorus (introduced at rate $L$), and the availability of light at the surface ($I_0$) and the density dependent growth response to phytoplankton in response to light ($\gamma$, Eq.
+Phytoplankton death is due to mortality and sinking at rate $m$ or predation by herbivores at rate $c_{HF}$, while their growth is controlled by the per capita growth rate ($r_P$), the availability of phosphorus (introduced at rate $L$), and the availability of light at the surface ($I_0$) and the density dependent growth response to phytoplankton in response to light ($\gamma$, Eq.
 5).
@@
 Plotting this equation using established ranges for ecological parameters (Table 1; @Carpenter2008; @Carpenter2013),  Figure \ref{fig4} shows that the system should support two alternative stable states, and can undergo a fold bifurcation (in reverse, creating a new stable point) when "catch" $q$ is decreased sufficiently.
-This sudden transition is felt in other dimensions as it's impact cascades down to the lower trophic levels.
+This sudden transition is felt in other dimensions as its impact cascades down to the lower trophic levels.
 
@@
 The small cumulative increase in $q$ leading up to step 20 caused a slight increase in $F$ and a dramatic decline in $H$ ($F$'s prey); after its initial decline, the abundance of $H$ was 0.001 for the remainder of the time series.
-When H abruptly declined, P abruptly increased, then continued to increased approximately linearly for the remainder of the time series.
+When $H$ abruptly declined, $P$ abruptly increased, then continued to increased approximately linearly for the remainder of the time series.
 Therefore, increasing $q$ caused declines in both bass states ($A$ and $J$), an increase in planktivorous fish ($F$), a decline in herbivorous zooplankton ($H$), and an increase in phytoplankton ($P$), a pattern consistent with a trophic cascade.
@@
 
-This analysis also highlights the importance of what might appear a merely semantic issue that in fact corresponds to fundamentally different dynamics.
+This analysis also highlights the importance of what might appear to be a merely semantic issue that in fact corresponds to fundamentally different dynamics.
 In this formulation, the control parameter $q$ corresponds to harvest (positive $q$) or stocking (negative $q$) of adult bass in the lake.
@@
 decreasing $q$, effectively running the scenario of increasing fishing harvests in reverse) gives the rest of the food web time to adjust.
-In contrast, a single large introduction of many bass would create an S-tipping transition, driving the system immediately into an alternate stable state without exhibiting CSD phenomena needed to evaluate early warning signals of a tipping point.
+In contrast, a single large introduction of many bass would create an S-tipping transition, driving the system immediately into an alternate state without exhibiting CSD phenomena needed to evaluate early warning signals of a tipping point.
 
@@
 Under a stochastic model, the distinction between S-tipping and B-tipping is slightly blurred, since a system that gets close enough to the tipping point due to changing parameters (B-tipping) will ultimately be able to jump across the separatrix (S-tipping) before the actual bifurcation event.
-In a stochastic system, fluctuations to the state variable act as intrinsic sources of perturbations, but though part of the system, large chance deviations are no more predictable than sudden exogenous forcing: there is no CSD prior to purely stochastic transitions [@PRSB2013], even though retrospectively a large deviation will share similar statistical signatures such as rapidly increasing variance and auto-correlation [@PRSB2012].
+In a stochastic system, fluctuations in the state variable act as intrinsic sources of perturbations, and though part of the system, large chance deviations are no more predictable than sudden exogenous forcing: there is no CSD prior to purely stochastic transitions [@PRSB2013], even though retrospectively a large deviation will share similar statistical signatures such as rapidly increasing variance and autocorrelation [@PRSB2012].
 
@@
 Influential reviews on abrupt change in ecological systems [e.g.
-@Ratajczak2018] focus almost exclusively on descriptions of one-dimensional state-spaces without discussion of if, when or how complex systems such as the ecosystem food-web considered here can be appropriately described by a non-linear equation in a single dimension.
-The analysis shown here illustrates this process on the basis of a fast-slow timescale approximation, which allows us to replace the faster dynamics at lower trophic levels into their equilibrium states.
-(Note also that the non-linearity necessary to support alternate stable states only emerges through this approximation.)  Thus, for the five-dimensional model considered here, our analysis shows that the familiar one-dimensional characterization of a sudden shift being driven by a fold bifurcation is indeed appropriate, but this was not all-together evident a priori.
+@Ratajczak2018] focus almost exclusively on descriptions of one-dimensional state-spaces without discussion of if, when or how complex systems such as the ecosystem food-web considered here can be appropriately described by a nonlinear equation in a single dimension.
+The analysis shown here illustrates this process on the basis of a fast-slow timescale approximation, which allows us to replace the faster dynamics at lower trophic levels with their equilibrium states.
+(Note also that the nonlinearity necessary to support alternate stable states only emerges through this approximation.)  Thus, for the five-dimensional model considered here, our analysis shows that the familiar one-dimensional characterization of a sudden shift being driven by a fold bifurcation is indeed appropriate, but this was not altogether evident a priori.
 The fact that CSD phenomena can occur in many different models has been suggested to mean that CSD is a generic or universal early warning indicator [@Scheffer2009] of a critical transition.

### 2018-07-15 paper/paper.Rmd
remove confusing previous sentence at end of intro

@@
 Common versions of this model include 5 state variables, ranging from top predators to primary producers (Figure \ref{fig2}).
-We perform dimension reduction on the full 5D model, and surmise whether the reduced model can be expressed in a way such that manipulating predator abundance is tantamount to manipulating a parameter, or if it necessarily constitutes state manipulation.
+We perform dimension reduction on the full 5D model.
 This exercise facilitates the interpretation of the lake experiment in the context of predicting regime shifts by using a 1D model to understand a high dimensional system.

### 2018-07-15 paper/paper.Rmd
fix subscript and rebuild pdf

@@
 3).
-Herbivorous zooplankton diffuse between surface waters and a deep water refuge (of size $H_R_$) at a rate $D_H$, grow by preying on phytoplankton at rate $c_{HP}$ and converting phytoplankton biomass (phosphorus) by a factor $\alpha$, and decline as they are preyed upon by planktivores at a rate $c_{HF}$ (Eq.
+Herbivorous zooplankton diffuse between surface waters and a deep water refuge (of size $H_R$) at a rate $D_H$, grow by preying on phytoplankton at rate $c_{HP}$ and converting phytoplankton biomass (phosphorus) by a factor $\alpha$, and decline as they are preyed upon by planktivores at a rate $c_{HF}$ (Eq.
 4).

### 2018-07-15 paper/paper.Rmd
re-render

@@
   - code: rb
-    address: "Rensselaer Polytechnic Institute, Department of Biological Sciences, Troy, NY 12309 USA; 2Rutgers University, Department of Ecology, Evolution, and Natural Resources, New Brunswick, NJ 08901 USA"
+    address: "Rensselaer Polytechnic Institute, Department of Biological Sciences, Troy, NY 12309 USA; Rutgers University, Department of Ecology, Evolution, and Natural Resources, New Brunswick, NJ 08901 USA"
 

### 2018-11-09 analysis/manuscript_report.Rmd
R -> Rmd in analyses

---
title: "Bifurcation or state tipping: assessing transition type in a model trophic cascade"
author: "Ryan Batt & Carl Boettiger"
date: "2018-07-07"
abstract: |
       The primary purpose of this document is to sketch out the key equations, figures, and numerical results for a manuscript that focuses on assessing whether a model undergoes state-tipping or a bifurcation. Our approach emphasizes reducing the dimensionality of the original model to redefine manipulated state variables as parameters. Thus, when these redefined parameters are manipulated to induce a transition, because they parameters not states, the transition should be viewed as a bifurcation, not state tipping.
output:
  # html_document:
  #   toc: true
  #   toc_depth: 4
  #   fig_caption: true
  #   theme: "readable"
  #   template: default
  pdf_document:
    toc: true
    toc_depth: 4
    keep_tex: true
#    template: latex-ryan.template
    fig_caption: yes
geometry: margin=1.0in
lineno: true
lineSpacing: false
titlesec: true
documentclass: article
placeins: true
---

```{r report-setup, include=FALSE, echo=FALSE, cache=FALSE}
# ==========================================
# = Record Time to Get Elapsed Time at End =
# ==========================================
t1 <- Sys.time()

# =================
# = Load Packages =
# =================
library(viridis)
library(phaseR)
library(rootSolve)
library(bs.tipping)

# Report
library(knitr)
library(rmarkdown)

# ================
# = Report Setup =
# ================
doc_type <- c("html", "pdf")[1]
table_type <- c("html" = "html", "pdf" = "latex")[doc_type]
options("digits" = 3) # rounding output to 4 in kable() (non-regression tables)
o_f <- paste(doc_type, "document", sep = "_")

opts_chunk$set(
  fig.path = "manuscript_report/",
  cache.path = "manuscript_report/",
  echo = TRUE,
  include = TRUE,
  cache = FALSE,
  autodep = TRUE,
  results = "asis",
  warning = FALSE,
  fig.show = "hold",
  fig.lp = NULL
)
```

# System of Equations & Manipulation

Start with the equations for the ecosystem dynamics:

$$\begin{array}{llr}
\dot{A} &= sJ - qEA - m_AA &(1) \\
\dot{F} &= D_F(F_o - F) - c_{FA}FA &(2)\\
\dot{J} &= fA - c_{JA}JA - \frac{c_{JF}vJF}{h+v+c_{JF}F} - sJ &(3)\\
\dot{H} &= D_H(H_R-H) + \alpha c_{HP}HP - c_{HF}HF &(4)\\
\dot{P} &= r_P L \gamma (I_0,P)P - mP - c_{PH}HP &(5)\\
\end{array}$$

Focusing on the fish, and setting each of the 3 fish differential equations to 0 and solving for the respective state variables:

$$\begin{array}{llr}
A&= \frac{sJ}{qE+m_A} &(6)\\[10pt]
F&=\frac{D_FF_0}{D_F + c_{FA}A} &(7)\\[10pt]
J&=\frac{fA}{c_{JA}A+s+\frac{c_{JF}vF}{h+v+c_{JF}F}} &(8)\\[10pt]
\end{array}$$

**Begin Equations of Motion**  
Although our ultimate goal is to see if the equation of motion for $J$ or $F$ shows alternate states, for completeness we will also present the equation of motion for $A$, which is acquired by substituting the RHS of Eq7 for $F$ in Eq8, then substituting the resulting expression for $J$ in Eq1:

$$\begin{array}{llr}
\dot{A} &=  sfA\Bigg(c_{JA}A+s+c_{JF}v\bigg(\frac{(h+v)(D_F+c_{FA}A)}{D_F F_o}+c_{JF}\bigg)^{-1}\Bigg)^{-1} -qA -m_A A &(9)
\end{array}$$

To get the equation of motion for $F$, we begin by getting an expression for $A$ that depends only on $F$ and parameters, which is achieved by substituting Eq8 into Eq6 and solving for $A$:

$$\begin{array}{llr}
A&=\bigg({\frac{sfA}{c_{JA}A+s+\frac{c_{JF}vF}{h+v+c_{JF}F}}}\bigg)\bigg(qE+m_A\bigg)^{-1} &(10a)\\[15pt]
sfA &= A(qE+m_A)\big(c_{JA}A+s+\frac{c_{JF}vF}{h+v+c_{JF}F}\big) &(10b)\\[10pt]
sf &=(qE+m_A)\big(c_{JA}A+s+\frac{c_{JF}vF}{h+v+c_{JF}F}\big) &(10c)\\[10pt]
A&=\frac{sf}{c_{JA}(qE+m_A)} - \frac{s}{c_{JA}} - \frac{c_{JF}vF}{c_{JA}(h+v+c_{JF}F)} &(10d)
\end{array}$$

Going from Eq10b to Eq10c is achieved by dividing both sides of the equation by $A$, which implies that in Eq10c $A$ cannot be 0 when $\dot{A}$ is 0. We will return to the consequences of this assumption later as we assess our derived equation of motion for $F$. By subtituting the RHS of Eq10d for $A$ in Eq2, we arrive at the equation of motion for $F$:

$$\begin{array}{llr}
\dot{F} &= D_F(F_o-F) - c_{FA}F(\frac{sf}{c_{JA}(qE+m_A)} - \frac{s}{c_{JA}} - \frac{c_{JF}vF}{c_{JA}(h+v+c_{JF}F)}) &(11)
\end{array}$$

Eq10 is what is used in `dF_dt_1state`.

The equation of motion for $J$ can be acquired by substituting the RHS of Eq7 for $F$ in Eq3 ,then exchanging $A$ in that resulting expression for the RHS of Eq6:

$$\begin{array}{llr}
Q &\equiv 1/(qE + m_A) \\
 \dot{J} &= fsJQ - c_{JA}sQJ^2 - sJ - \frac{c_{JF}vJ}{\frac{(h+v)}{(D_FF_o)(D_F+c_{FA}sQJ)^{-1}} + c_{JF}} &(12)
\end{array}$$

***

# Table of Parameter Values

```{r parameterTable, results='markup', echo=FALSE}

# Symbol
# Description
# Quantity
# Units
paramDefs <- c(
  # "A",
  # "Adult bass",
  # "variable",
  # "kg ha^{-1}", # so, it's actually number of fish, but if you multiply by 0.2 kg/fish you get kg/ha
  #
  # "F",
  # "Planktivorous fish",
  # "variable",
  # "kg ha^{-1}", # actually n, but divide by 500 to get kg/fish
  #
  # "J",
  # "Juvenile bass",
  # "variable",
  # "kg ha^{-1}", # actually n, but divide by 10 to get kg/ha
  #
  # "H",
  # "Herbivorous zooplankton",
  # "variable",
  # "", # no biomass conversion listed in carpenter et al 2008 table
  #
  # "P",
  # "Phytoplankton",
  # "variable",
  # "", # no biomass conversion listed in carpenter et al 2008 table

  "$q$",
  "Harvest or stocking",
  "variable",
  "$t^{-1}$",

  "$s$",
  "Survival rate of juveniles to maturation",
  "0.6",
  "$t^{-1}$",

  "$m_A$",
  "Mortality rate of adults",
  "0.4",
  "$t^{-1}$",

  "$m$",
  "Phytoplankton mortality",
  "0.1",
  "$t^{-1}$",

  "$f$",
  "Fecundity of adult bass",
  "2",
  "$J/A$",

  "$c_{FA}$",
  "Predation of planktivores by adult bass",
  "0.3",
  "$t^{-1} A^{-1}$",

  "$c_{JA}$",
  "Predation of juvenile bass on by adult bass",
  "0.1",
  "$t^{-1} A^{-1}$",

  "$c_{JF}$",
  "Predation of juvenile bass by planktivorous fish",
  "0.5",
  "$t^{-1} F^{-1}$",

  "$c_{PH}$",
  "Predation of phytoplankton by herbivores",
  "0.25",
  "$t^{-1} H^{-1}$",

  "$c_{HF}$",
  "Predation of herbivores by planktivores",
  "0.1",
  "$t^{-1} F^{-1}$",

  "$F_o$",
  "Abundance of planktivorous fish in non-foraging arena",
  "200",
  "F",

  "$H_R$",
  "Reservoir of herbivores in a deep-water refuge",
  "4",
  "H",

  "$D_F$",
  "Diffusion of planktivores between refuge and foraging arena",
  "0.09",
  "$t^{-1}$",

  "$DH$",
  "Diffusion of between shallow and deep water",
  "0.5",
  "$t^{-1}$",

  "$v$",
  "Rate at which J enter a foraging arena, becoming vulnerable",
  "80",
  "$t^{-1}$",

  "$h$",
  "Rate at which J hide in a refuge",
  "80",
  "$t^{-1}$",

  "$r_P$",
  "Growth rate of phytoplankton",
  "3",
  "$m^2 mg^{-1}$",

  "$L$",
  "Phosphorus loading rate",
  "0.6",
  "$mg m^{-2} t^{-1}$",

  "$\\alpha$",
  "Assimilation of phytoplankton phosphorus by zooplankton",
  "0.3",
  "$g g^{-1}$",

  "$\\gamma$",
  "Density-dependent phytoplankton growth response to light",
  "n/a",
  "n/a",

  "$I0$",
  "Solar irradiance incident on the surface of the water",
  "300",
  "$\\mu E m^{-2} s^{-1}$"
)
colNames <- c("Symbol", "Description", "Quantity", "Units")
paramTable_mat <- matrix(paramDefs, byrow = TRUE, ncol = 4, dimnames = list(NULL, colNames))
kable(paramTable_mat, caption = "**Table 1.** Model parameters, description, values, and units. Parameters are taken from Carpenter et al (2008) and Carpenter et al (2013). The function $\\gamma$ describes phytoplankton growth, and can be found in Carpenter et al (2008).")
```

# Setup

```{r setup}
set.seed(42)
qE <- -0.001 # 0.02
qE_end <- 0.001 # /20 #-5#-0.025

dt <- 0.01 # 0.01
nYears <- 1020 # /20

noise_coeff <- c(0.01, 0.01, 0.01, 0.01, 0.01)
```

# Simulation
## Simulate Time series

```{r simulation}
# set initial values to equilibrium (roots)
X_init <- pmax(bs.tipping::getRoot5D(c(A0 = 150, F0 = 50, J0 = 100, H0 = 100, P0 = 500),
                                     pars = c(qE = qE)), 1E-3)

# simulate an example of fish
qE_vec <- seq(qE, qE_end, length.out = nYears / dt)
stateMat <- matrix(NA, nrow = nYears / dt, ncol = 5, 
                   dimnames = list(NULL, c("A0", "F0", "J0", "H0", "P0")))
stateMat[1, ] <- unlist(X_init)

for (j in 2:nrow(stateMat)) { # iterate through time steps
  state <- stateMat[j - 1, ]
  dState_dt <- ecoStep(X = state, pars = c(qE = (qE_vec[j])))

  # Euler Method Approximation
  dState <- dState_dt * dt + rnorm(5, sd = c(noise_coeff)) * dt
  eulerState <- pmax(state + dState, 1E-3)
  stateMat[j, ] <- eulerState # euler
  # stateMat[j,] <- getRoot5D(stateMat[j-1,], pars=c(qE=qE_vec[j]))
}

tsMat <- cbind(time = seq(0, nYears - dt, by = dt), qE = qE_vec, stateMat)
```

## Figure: 5-dimension simulation

```{r figure-5dsimulation, fig.width=3.5, fig.height=6, fig.cap="Simulated time series of adult bass (A0), planktivores (F0), and juvenile bass (J0) over a gradient of harvesting of adult bass (qE)."}
par(mfrow = c(5, 1), mar = c(1.5, 2, 0.75, 0.25), ps = 8, 
    cex = 1, mgp = c(1, 0.25, 0), tcl = -0.15)
stateLabs <- c(
  A0 = "A (Adult Bass)",
  F0 = "F (Planktivores)",
  J0 = "J (Juvenile Bass)",
  H0 = "H ( Herbivores)",
  P0 = "P (Phytoplankton)"
)
stateNames <- names(stateLabs)
for (i in 1:length(stateNames)) {
  plot(tsMat[, "time"], tsMat[, stateNames[i]], type = "l",
       ylab = stateLabs[stateNames[i]], xlab = "")
}
```

***

# Equations of Motion
## Figure: A Motion

```{r figure-eqMotion-A, fig.width=3.5, fig.height=3.5, fig.cap="Equation of motion for A."}
qevals <- c(0.25, 0.025, 0, -0.025)
par(mfrow = c(2, 2), mar = c(1.85, 1.85, 0.75, 0.25),
    mgp = c(1, 0.15, 0), tcl = -0.15, ps = 9, cex = 1)
for (i in 1:length(qevals)) {
  curve(stateMotion(state0 = x, pars = c(qE = qevals[i])),
        from = 0, to = 20, ylab = "dA/dt", xlab = "A")
  abline(h = 0, lty = 2)
  mtext(paste0("qE=", round(qevals[i], 2)),
        side = 3, line = 0, adj = 0.05, font = 2)
}
```

Looks good, the second stable node appears when qE>0

***

## Figure: J Motion

```{r figure-eqMotion-J, fig.width=3.5, fig.height=3.5, fig.cap="Equation of motion for J."}
par(mfrow = c(2, 2), mar = c(1.85, 1.85, 0.75, 0.25),
    mgp = c(1, 0.15, 0), tcl = -0.15, ps = 9, cex = 1)
for (i in 1:length(qevals)) {
  curve(stateMotion(state0 = x, stateVar = "J0", 
                    pars = c(qE = qevals[i])), 
        from = 0, to = 20, ylab = "dJ/dt", xlab = "J")
  abline(h = 0, lty = 2)
  mtext(paste0("qE=", round(qevals[i], 2)), 
        side = 3, line = 0, adj = 0.05, font = 2)
}
```

As with the adults, the second stable node appears when qE>0

***

## Figure: F Motion

```{r figure-eqMotion-F, fig.width=3.5, fig.height=3.5, fig.cap="Equation of motion for F."}
par(mfrow = c(2, 2), mar = c(1.85, 1.85, 0.75, 0.25),
    mgp = c(1, 0.15, 0), tcl = -0.15, ps = 9, cex = 1)
for (i in 1:length(qevals)) {
  curve(stateMotion(state0 = x, stateVar = "F0", pars = c(qE = qevals[i])),
        from = 0, to = 20, ylab = "dF/dt", xlab = "F")
  abline(h = 0, lty = 2)
  mtext(paste0("qE=", round(qevals[i], 2)), 
        side = 3, line = 0, adj = 0.05, font = 2)
}
```

Unfortunately, this state variable only shows a trivial stable node and a saddle point. This is the state variable that requires algebra cancelling the 'A' state variable in order to get its equation of motion (without A canceling, you cannot get a dF/dt vs F type plot easily [as a function of F and parameters]). We think this result is invalid, as there must be some assumptions in the algebra that invalidate our desired interpretation.

***

## Figure: Combined A, J, F Motion

```{r figure-eqMotion-AJF, fig.width=5, fig.height=5, fig.cap="Motion of A, J, and F. Note that the equation of motion for F failed to exhibit alternate states. In formulating the equation for motion of F, there is an algebraic step that requires dropping from terms via division by A, which assume that $A!=0$. Therefore, states of F that correspond to $A=0$ are invalid. The equation of motion for A indicates that one of the stable nodes is at $A=0$, implying that the equation of motion for F is inaccurate in the corresponding region."}
xlabs <- c("A", "J", "F") # ylabs <- c("dA/dt", "dJ/dt", "dF/dt")
ylabs <- paste0("d", xlabs, "/dt")
svars <- paste0(xlabs, "0")
par(mfrow = c(3, 3), mar = c(1.85, 1.85, 0.75, 0.25), 
    mgp = c(1, 0.15, 0), tcl = -0.15, ps = 9, cex = 1)
for (s in 1:3) {
  for (i in 2:length(qevals)) {
    curve(stateMotion(state0 = x, stateVar = svars[s], pars = c(qE = qevals[i])),
          from = 0, to = 20, ylab = ylabs[s], xlab = xlabs[s])
    abline(h = 0, lty = 2)
    mtext(paste0("qE=", round(qevals[i], 2)), side = 3, line = 0, adj = 0.05, font = 2)
  }
}
```

***

## Figure: A Motion A Ball-in-Cup

```{r}
Adot <- function(A, pars = c(qE = -0.001)) {
  parsF <- unlist(
    formals(ecoStep)[c("fA", "cJA", "cJF", "cFA", 
                       "vuln", "hide", "surv", "Fo", "DF")])
  if (missing(pars)) { 
    # if a function requires qE, pars needs to be supplied (e.g., pars=c(qE=0.1))
    pars <- parsF
  } else {
    pars <- c(pars, parsF[!names(parsF) %in% names(pars)])
  }
  with(as.list(pars), {
    surv * fA * A * (cJA * A + surv + cJF * vuln * 
                       (((hide + vuln) * (DF + cFA * A)) / 
                          (DF * Fo) + cJF)^(-1))^(-1) - 
      qE * A - (mA <- 0.4) * A
  })
}
```

```{r include = FALSE, eval=FALSE}
# Avec <- seq(0, 20, length.out=200)
# dAdt_A_vec <- Adot(A=Avec, pars=c(qE=-0.001))
# plot(Avec, dAdt_A_vec, type='l')
# # curve(Adot(A=x, pars=c(qE=-0.001)), from=0, to=20); abline(h=0)
# plot(Avec, cumsum(dAdt_A_vec), type='l')

# Avec <- seq(0, 20, length.out=200)
# qevals2 <- c(qevals[-1], -1)
# par(mfcol=c(2,4), mar=c(1.85,1.85,0.75,0.25), mgp=c(1,0.15,0), tcl=-0.15, ps=9, cex=1)
# l <- 0
# for(q in 1:length(qevals2)){
# 	dAdt_A_vec <- Adot(A=Avec, pars=c(qE=qevals2[q]))
# 	par(mar=c(1.85,1.85,0.75,0.25))
# 	plot(Avec, dAdt_A_vec, type='l', ylab="dA/dt", xlab="A")
# 	qlab <- paste0("q=",round(qevals2[q],2))
# 	mtext(qlab, side=3, line=0, adj=0.05, font=2)
# 	abline(h=0, lty=2)
# 	mtext(LETTERS[l<-l+1], side=3, line=-1, adj=0.05)
#
# 	par(mar=c(1.0,1.85,1.6,0.25))
# 	plot(Avec, -cumsum(dAdt_A_vec), type='l', ylab="", xlab="", xaxt='n', yaxt='n')
# 	mtext(qlab, side=3, line=0, adj=0.05, font=2)
# 	mtext("A", side=1, line=0)
# 	mtext(LETTERS[l<-l+1], side=3, line=-1, adj=0.05)
# }
```

```{r figure-Amotion-ballincup, fig.width=6, fig.height=3.5, fig.cap="Motion of A, and its integral (ball-in-cup diagram)."}
Avec <- seq(-0.75, 20, length.out = 200)
qevals2 <- c(qevals[-1], -0.5)
par(mfrow = c(2, 4), mar = c(1.85, 1.85, 0.75, 0.25), 
    mgp = c(1, 0.15, 0), tcl = -0.15, ps = 9, cex = 1)
l <- 0
for (q in 1:length(qevals2)) {
  dAdt_A_vec <- Adot(A = Avec, pars = c(qE = qevals2[q]))
  qlab <- paste0("q=", round(qevals2[q], 2))
  # par(mar=c(1.85,1.85,0.75,0.25))
  ylim <- range(dAdt_A_vec)
  if (ylim[2] < 0.5) {
    yadj <- diff(ylim) * 0.15
    ylim[2] <- ylim[2] + yadj # max(ylim[2], 0.15)
  }
  plot(Avec, dAdt_A_vec, type = "l", ylab = "dA/dt", xlab = "A", ylim = ylim)
  mtext(qlab, side = 3, line = 0, adj = 0.05, font = 2)
  abline(h = 0, lty = 2)
  mtext(LETTERS[l <- l + 1], side = 3, line = -0.8, adj = 0.025)
}
for (q in 1:length(qevals2)) {
  dAdt_A_vec <- Adot(A = Avec, pars = c(qE = qevals2[q]))
  qlab <- paste0("q=", round(qevals2[q], 2))
  # par(mar=c(1.0,1.85,1.6,0.25))
  plot(Avec, -cumsum(dAdt_A_vec), type = "l", 
       ylab = "", xlab = "A", xaxt = "s", yaxt = "n")
  mtext(qlab, side = 3, line = 0, adj = 0.05, font = 2)
  # mtext("A", side=1, line=0)
  mtext(LETTERS[l <- l + 1], side = 3, line = -1, adj = 0.025)
}
```

***

# Figure: Bifurcation Diagram

```{r }
qe_range <- c(0.075, -0.1) # range of q values
qevec_bd <- do.call(seq, c(as.list(qe_range), length = 2E3)) # sequence of q values
rts <- vector("list", length = length(qevec_bd)) # empty list to store roots for each q value
for (q in 1:length(qevec_bd)) { # for each val of q
  tq <- qevec_bd[q] # temp q value
  trts <- rootSolve::uniroot.all(f = Adot, interval = c(0, 20),
                                 pars = c(qE = tq)) # temp roots
  rts[[q]] <- trts # store temp root
  tclasses <- vector("character", length = length(trts))
  for (r in 1:length(trts)) { # for each root found for this val of q
    trt <- trts[r] # temp root
    t_perturb <- max(0.01 * trt, 0.01)
    resp_neg <- Adot(trt - t_perturb, pars = c(qE = tq))
    resp_pos <- Adot(trt + t_perturb, pars = c(qE = tq))
    if (resp_neg > 0 & resp_pos < 0) { # crosses 0 line downward
      tclass <- "snode"
    } else if (sign(resp_neg) == sign(resp_pos)) { # barely touches 0 line
      tclass <- "unode"
    } else if (resp_neg < 0 & resp_pos > 0) { 
      # cross 0 line upward; 
      tclass <- "unode"
    }
    tclasses[r] <- tclass
  }
  names(rts[[q]]) <- tclasses
}
```

```{r}
# names(rts) <- paste0('q',round(qevec_bd,4))
get_bb <- function(x) min(x[names(x) %in% "snode"])
get_ub <- function(x) {
  snode_ind <- names(x) %in% "snode"
  if (sum(snode_ind) > 1) {
    return(unname(max(x[snode_ind])))
  } else {
    return(NA)
  }
}
```

```{r}
get_separatrix <- function(x) {
  unode_ind <- names(x) %in% "unode"
  if (any(unode_ind)) {
    return(unname(x[unode_ind]))
  } else {
    return(NA)
  }
}
```

```{r figure-bifurcationDiagram, fig.width=3.5, fig.height=3.5, fig.cap="Bifurcation diagram. Solid line is a stable state. Dashed line is an unstable state."}
bot_branch <- sapply(rts, get_bb)
upper_branch <- sapply(rts, get_ub)
separatrix <- sapply(rts, get_separatrix)
ylim <- range(c(bot_branch, upper_branch, separatrix), na.rm = TRUE)
# xlim <- rev(range(qevec_bd))
# plot(qevec_bd, bot_branch, ylim=ylim, xlim=xlim, type='l')
par(mar = c(2, 1.85, 0.75, 0.25), mgp = c(1, 0.15, 0), tcl = -0.15, ps = 9, cex = 1)
plot(qevec_bd, bot_branch, ylim = ylim, type = "l", 
     ylab = "A (adult bass)", xlab = "q (stocking or harvest)")
lines(qevec_bd, upper_branch, lty = 1)
lines(qevec_bd, separatrix, lty = 2)
```

***

# Session Info

```{r sessionInfo, results='markup'}
difftime(Sys.time(), t1) # how long it took to run these models/ produce this report
Sys.time()
sessionInfo()
```

### 2018-11-09 analysis/stability_report.Rmd
R -> Rmd in analyses

---
title: "Stability analysis of a trophic cascade model"
author: "Ryan Batt & Carl Boettiger"
date: "2018-06-16"
abstract: |
       Makings of a stability analysis of a trophic cascade model. Starting by focusing on fish (adult bass, juvenile bass, and a planktivore), though the model is easily extensible to include phytoplankton and zooplankton for a 5D model.
output:
  html_document:
    toc: true
    toc_depth: 4
    fig_caption: true
    theme: "readable"
    template: default
  pdf_document:
    toc: true
    toc_depth: 4
    fig_caption: yes
geometry: margin=1.0in
lineno: true
lineSpacing: false
titlesec: true
documentclass: article
placeins: true
---

```{r report-setup, include=FALSE, echo=FALSE, cache=FALSE}
# ==========================================
# = Record Time to Get Elapsed Time at End =
# ==========================================
t1 <- Sys.time()

# =================
# = Load Packages =
# =================
library(viridis)
library(phaseR)
library(rootSolve)
library(bs.tipping)

# Report
library(knitr)
library(rmarkdown)

# ================
# = Report Setup =
# ================
doc_type <- c("html", "pdf")[1]
table_type <- c("html"="html", "pdf"="latex")[doc_type]
options("digits"=3) # rounding output to 4 in kable() (non-regression tables)
o_f <- paste(doc_type, "document", sep="_")

opts_chunk$set(
	fig.path = 'stability_report/', 
	cache.path='stability_report/',
	echo=TRUE, 
	include=TRUE, 
	cache=FALSE,
	autodep=TRUE,
	results='asis',
	warning=FALSE,
	fig.show="hold"
)
```

#Setup

```{r setup}
qE <- 0.65
qE_end <- 0.1

dt <- 0.01
nYears <- 1000

noise_coeff <- c(0.01, 0.01, 0.01)
```

#Simulation

```{r simulation}

# set initial values to equilibrium (roots)
X_init <- pmax(bs.tipping::getRoot(c(A0=1, F0=1, J0=1), pars=c(qE=qE)), 1E-2)

# simulate an example of fish
qE_vec <- seq(qE, qE_end, length.out=nYears/dt)
stateMat <- matrix(NA, nrow=nYears/dt, ncol=3, dimnames=list(NULL, c("A0","F0", "J0")))
stateMat[1,] <- unlist(X_init)

for(j in 2:nrow(stateMat)){ # iterate through time steps
	state <- stateMat[j-1,]
	dState_dt <- fishStep(X=state, pars=c(qE=(qE_vec[j])))

	# Euler Method Approximation
	dState <- dState_dt*dt + rnorm(3, sd=c(noise_coeff[1], noise_coeff[2], noise_coeff[3]))*dt 
	eulerState <- pmax(state + dState, 1E-2)
	stateMat[j,] <- eulerState # euler
}

stateMat <- cbind(time=seq(0, nYears-dt, by=dt), qE=qE_vec, stateMat)
```

## Plot Simulation

```{r plot-simulation, fig.width=3.5, fig.height=6, fig.cap="Simulated time series of adult bass (A0), planktivores (F0), and juvenile bass (J0) over a gradient of harvesting of adult bass (qE)."}
par(mfrow=c(3,1), mar=c(2, 2, 0.75, 0.25), ps=8, cex=1, mgp=c(1, 0.25, 0), tcl=-0.15)
plot(stateMat[,c("time","A0")], type='l')
plot(stateMat[,c("time","F0")], type='l')
plot(stateMat[,c("time","J0")], type='l')
```

***

## Rearrange equations to represent stability in 1D

Start with the equations for the fish dynamics:

$$\begin{array}{llr}
\dot{A} &= sJ - qEA - (1-s)A &(1) \\
\dot{F} &= d(F_o - F) - c_{FA}FA &(2)\\
\dot{J} &= fA - c_{JA}JA - \frac{c_{JF}vJF}{h+v+c_{JF}F} - sJ &(3)
\end{array}$$

Then, set $\dot{A}$ and $\dot{J}$ to 0, and solve for $A$ and $J$, respectively:  
$$\begin{array}{llr}
A&= \frac{sJ}{qE+1-s} &(4)\\[10pt]
J&=\frac{fA}{xA+s+\frac{zvF}{h+v+zF}} &(5)\\[10pt]
\end{array}$$

Substitute Eq5 into Eq4, and solve for $A$:  
$$\begin{array}{llr}
A&=\frac{sf}{x(qE+1-s)} - \frac{s}{x} - \frac{zvF}{x(h+v+zF)} &(6)
\end{array}$$

Substitute Eq6 into Eq2, giving us the dynamics of $F$ as a function of $F$ and parameters:  
$$\begin{array}{llr}
\dot{F} &= d(F_o-F) - aF(\frac{sf}{x(qE+1-s)} - \frac{s}{x} - \frac{zvF}{x(h+v+zF)}) &(7)
\end{array}$$

Eq7 is what is used in `dF_dt_1state`.

**Alternatively**, we can rearrange the equations to perform a different seat of substituions to solve for $\dot{J}$ as a function of $J$ and parameters (juvenile bass):  
$$\begin{array}{llr}
Q &\equiv 1/(qE + 1 - s) \\
 \dot{J} &= fsJQ - xsQJ^2 - sJ - \frac{zvJ}{\frac{(h+v)}{(dF_o)(d+asQJ)^{-1}} + z} &(8)
\end{array}$$

***

# Plots of dState/dt vs State

## dF/dt vs F

```{r plot-dFdt-vs-F, fig.width=3.5, fig.height=3.5, fig.cap="dF/dt vs F. Where the line intersects 0, indicates a value of planktivore abundance that is an equilibrium. Is contingent upon the value of parameters, such as qE. The value of qE used here is the value used for the first time step of the simulation."}
# check rate equation ... should be 0 at root
F_grad <- seq(-2, 110, length.out=100)
par(mar=c(2,2,0.75,0.5), mgp=c(1,0.25,0), tcl=-0.15, ps=8, cex=1)
plot(F_grad, dFJ_dt_1state(State0=F_grad, pars=c(qE=qE), stateName="F0"), type='l', ylab="dF/dt", xlab='F')
mtext(paste0('qE = ', round(qE,2)), line=-0.1, adj=0.05, font=2)
abline(h=0, lty=2)
```

This plot of the potential looks wrong to me. At a harvest rate of qE=0.65, F should have an equilibrium near F=100, but this plot shows a quadratic where the change in F is getting more and more positive past ~10. Furthermore, 10 looks like a saddle, not a stable node.

## dJ/dt vs J

```{r plot-dJdt-vs-J, fig.width=3.5, fig.height=3.5, fig.cap="dJ/dt vs J Where the line intersects 0, indicates a value of juvenile bass abundance that is an equilibrium. Is contingent upon the value of parameters, such as qE. The value of qE used here is the value used for the first time step of the simulation."}
# check rate equation ... should be 0 at root
J_grad <- seq(0, 850, length.out=1E3)
par(mar=c(2,2,0.75,0.5), mgp=c(1,0.25,0), tcl=-0.15, ps=8, cex=1)
plot(J_grad, dFJ_dt_1state(State0=J_grad, pars=c(qE=qE), stateName="J0"), type='l', ylab="dJ/dt", xlab='J')
mtext(paste0('qE = ', round(qE,2)), line=-0.1, adj=0.05, font=2)
abline(h=0, lty=2)
```

## dJ/dt vs J for various qE

```{r plot-dJdt-vs-J-qE, fig.width=3.5, fig.height=3.5, fig.cap="dJ/dt vs J Where the line intersects 0, indicates a value of juvenile bass abundance that is an equilibrium. Is contingent upon the value of parameters, such as qE. Thus, the dJ/dt vs J curve is plotted for several values of qE, providing a visualization of how the stability landscape changes with harvest rate."}
# check rate equation ... should be 0 at root
J_grad <- seq(-1, 850, length.out=1E3)
# qE_vals <- seq(qE, qE_end, length.out=4)
qE_vals <- seq(1.25, 0.25, length.out=8)
dFJ_dt_1state_Jwrap <- function(X){dFJ_dt_1state(State0=J_grad, pars=c(qE=X), stateName="J0")}
dJ_dt_qE <- lapply(qE_vals, FUN=dFJ_dt_1state_Jwrap)
names(dJ_dt_qE) <- paste0("qE",round(qE_vals,2))

par(mar=c(2,2,0.75,0.5), mgp=c(1,0.25,0), tcl=-0.15, ps=8, cex=1)
ylim <- range(dJ_dt_qE)
cols <- viridis(n=length(qE_vals))
plot(J_grad, dJ_dt_qE[[1]], type='l', ylab="dJ/dt", xlab='J', ylim=ylim, col=cols[1])
for(j in 2:length(qE_vals)){
	lines(J_grad, dJ_dt_qE[[j]], col=cols[j])
}
legend('bottomleft', lty=1, col=cols, legend=paste0("qE = ",round(qE_vals,2)), ncol=2, y.intersp=0.6, x.intersp=0.5)
abline(h=0, lty=2)
```

***

# Sanity Check on dF/dt

## Part 1
The code below is taken from the examples in the help file for `dF_dt_1state`.

```{r sanity-check1, results='markup'}
getRoot(c(A0=1000, F0=1, J0=1000), pars=c(qE=0.5)) # find equilibria when bass are abundant
dFJ_dt_1state(State0=0.06712064, pars=c(qE=0.5), stateName="F0") # F0 set to equilibrium when bass start as abundant
```

Check -- results make sense, the `getRoot` function finds that when bass start off super abundant, planktivores should stabilize near `0`. Using the `dF_dt_1state` function, we find that the change in planktivore abundance per unit time is very near `0` when planktivore abundance is near `0`. Great.

## Part 2

```{r sanity-check2, results='markup'}
getRoot(c(A0=1, F0=1, J0=1), pars=c(qE=0.5)) # find equilibria when bass are rare
dFJ_dt_1state(State0=17.890184, pars=c(qE=0.5), stateName='F0') # F0 set to equilibrium when bass start as rare
```

Check -- again, we find that according to `dF_dt_1state`, $dF/dt$ is near `0` when fish abundance is set near to equilibrium. This time the equilibrium value for planktivores was higher because the bass starting point was low. Right. Great.

## Part 3

```{r sanity-check3, results='markup'}
getRoot(c(A0=1000, F0=1, J0=1000), pars=c(qE=0.65)) # find equilibria when bass are abundant
dFJ_dt_1state(State0=0.09136212, pars=c(qE=0.65), stateName="F0") # check planktivore equation for rate of change
fishStep(X=c(A0=364.51517642, F0=0.09136212, J0=838.38490576)) # re-examine rates of change of fish near equilibrium
```

Here we are repeating **Part 1**, but we've increased `qE` from `0.5` to `0.65`. I'm also cross-checking the dF/dt values from `dF_dt_1state` with those reported by `fishStep`, which is what `getRoot` uses to find equilibria. With the slightly higher harvest and high starting values for bass, we again find that planktivores should stabilize near `0`, and both `dF_dt_1state` and `fishStep` indicate that dF/dt is very small when setting F to this near-`0` equilibrium value. It's a little annoying that these two functions don't report the *exact* same value for dF/dt, but in general, this seems to check out. So far, things make sense.

## Part 4

```{r sanity-check4, results='markup'}
getRoot(c(A0=1, F0=1, J0=1), pars=c(qE=0.65)) # find equilibria when bass are rare
dFJ_dt_1state(State0=100, pars=c(qE=0.65), stateName="F0") # F0 set to equilibrium when bass start as rare # WTF?!
fishStep(X=c(A0=6.275129e-18, F0=100, J0=1.443280e-17)) # but this one says that dF/dt should be ~0!!
```

Here we again use the `qE=0.65` harvest rate, with low starting values for bass. But instead of `dF_dt_1state` reporting a near-`0` value for the rate of change in planktivores, it's reporting a very large positive value. This last example is what is making me scratch my head.

Update: apparently **Part 4** (above) doesn't work properly because in the algebra we dvide by A in a place that implies the assumption that $A != 0$. So when we try to use the equation when A=0, it doesn't work well.

***

# Stability Analysis

## Critical Values

```{r critVals}
(critVals <- findCritHarv()) # 0.24 and 1.22
```

These critical values were found numerically using the 1D model.

##Stability Classification

```{r stabClass, results="markup", eval = FALSE}
qEvec <- c(0, critVals[1]-0.2, critVals[1], critVals[2]-0.1, critVals[2]-0.05, critVals[2])
lout <- lapply(qEvec, stabClass)
do.call(rbind, lout)
```

<!--
I'm a little confused, because there is no place I could find with 2 stable nodes and 1 saddle point.  
-->

## Phase Portrait

```{r figure-phasePortrait, fig.width=6, fig.height=6, fig.cap="**Figure 1.** A phase portrait of the system for varying values of harvest (qE). The vector field (indicating the direction and speed that the system moves through phase space at that point) is represented by gray arrows. Nullclines are represented red and blue lines, indicating where dJ/dt and dF/dt are equal to zero, respectively.  Trajectories starting at arbitrary initial points (open diamonds) and continuing the along the accompanying solid black line indicate how the system moves from the initial point through phase space for 20 years. Equilibria are indicated by points: solid filled circle is a stable node, an 'X' is a saddle point. An equilibrium occurs whereever the nullclines cross. The different panels correspond to different values of harvest (qE). "}
qEvals <- rev(c(critVals[1]-0.2, critVals[1], critVals[2]-0.1, critVals[2]))
par(mfrow=c(2,2), mar=c(2,2,1,0.5), mgp=c(1,0.25,0), tcl=-0.15, cex=1, ps=9, cex.axis=0.85)
for(j in 1:4){
	(bs.tipping::phasePortrait(qE=qEvals[j], pars=NULL, nFlow=10, nNull=300, t.end=20, addLeg=TRUE))
	mtext(paste0("qE = ",round(qEvals[j],2)), side=3, line=0, adj=0, font=2)
}
```

***

## Growth and Consumption Curves: dJ/dt vs J

```{r figure-grow-cons-J, fig.width=6, fig.height=6, fig.cap="**Figure.** Growth and consumption curves for juvenile bass (J). Different panels show the curves for varying values of harvest rate on adult bass (qE)."}
# dFJ_dt_1state <- function(State0, pars, stateName=c("J0","F0"), parts=FALSE){
plot_growCons <- function(stateRange=c(0,999), pars=c(qE=1.0), stateName=c("J0","F0"), nGrid=100){
	stateName <- match.arg(stateName)
	stopifnot(length(stateRange)==2)

	names(stateRange) <- c("from", "to")
	grid_args <- c(as.list(stateRange), length.out=nGrid)
	stateGrid <- do.call(seq, grid_args)

	rates <- t(sapply(stateGrid, dFJ_dt_1state, pars=pars, stateName=stateName, parts=TRUE))
	state_rates <- data.table(state=stateGrid, rates)

	ylim <- state_rates[,range(c(growth, consumption))] #range(state_rates[,c("growth", "consumption")])
	state_rates[,plot(state,growth, col='blue', type='l', ylim=ylim, xlab="", ylab="")]
	state_rates[,lines(state,consumption, col='red')]

}
qEvals <- rev(c(critVals[1]-0.2, critVals[1], critVals[2]-0.1, critVals[2]))
par(mfrow=c(2,2), mar=c(2,2,1,0.5), mgp=c(1,0.25,0), tcl=-0.15, cex=1, ps=9, cex.axis=0.85)
for(j in 1:4){

	plot_growCons(pars=c(qE=qEvals[j]))

	mtext(paste0("qE = ",round(qEvals[j],2)), side=3, line=0, adj=0, font=2)
	mtext("dJ/dt", side=2, line=0.85)
	mtext("J", side=1, line=0.85)
	legend("topleft", legend=c("growth", "consumption"), col=c("blue","red"), lty=1)
}
```

***

# Session Info

```{r sessionInfo, results='markup'}
difftime(Sys.time(), t1) # how long it took to run these models/ produce this report
Sys.time()
sessionInfo()
```

### 2018-11-09 R/stability.R

@@ -251,9 +251,10 @@ stabClass <- function(qE, pars, func=fs2D, Jvals=seq(0,2E3,length.out=2), Fvals=
 	urs <- rs[!duplicated(paste0(round(rs[,"J0"],4),round(rs[,"F0"],4))), c("qE","J0","F0"), drop=FALSE]
 
 	stab <- function(x){
-		st <- phaseR::stability(fs2D, y.star=x[-1,drop=FALSE], parameters=c(qE=qE,pars), summary=FALSE)
+		st <- phaseR::stability(fs2D, ystar=x[-1,drop=FALSE], parameters=c(qE=qE,pars), summary=FALSE)
 		# if(is.null(names(st$parameters))){names(st$parameters) <- "I"}
-		o <- cbind(data.frame(J0=st$y.star[1], F0=st$y.star[2], classification=st$classification, Delta=st$Delta, discriminant=st$discriminant, tr=st$tr),as.list(st$parameters))
+		o <- cbind(data.frame(J0=st$ystar[1], F0=st$ystar[2], classification=st$classification,
+		                      Delta=st$Delta, discriminant=st$discriminant, tr=st$tr),as.list(st$parameters))
 		rownames(o) <- NULL
 		o$classification <- as.character(o$classification)
 		return(o)

### 2018-11-09 pkgBuild/dependencies.R

@@ -1,14 +1,0 @@
-
-update_dependencies <- function(){
-	devtools::use_package("rmarkdown", type="Imports")
-	devtools::use_package("testthat", type="Imports")
-	devtools::use_package("lintr", type="Imports")
-	devtools::use_package("rticles", type="Imports")
-	devtools::use_package("tidyverse", type="Suggests")
-	
-	devtools::use_package("stats", type="Imports")
-	devtools::use_package("methods", type="Imports")
-	devtools::use_package("rootSolve", type="Imports")
-	devtools::use_package("deSolve", type="Imports")
-	devtools::use_package("phaseR", type="Imports")
-}

### 2018-11-13 analysis/stability_report.Rmd
run stability_report as Rmd.

@@
 library(rmarkdown)
+library(data.table)
 
@@
 
-```{r figure-phasePortrait, fig.width=6, fig.height=6, fig.cap="**Figure 1.** A phase portrait of the system for varying values of harvest (qE).
+```{r figure-phasePortrait, eval=FALSE, fig.width=6, fig.height=6, fig.cap="**Figure 1.** A phase portrait of the system for varying values of harvest (qE).
 The vector field (indicating the direction and speed that the system moves through phase space at that point) is represented by gray arrows.
@@
 
-```{r figure-grow-cons-J, fig.width=6, fig.height=6, fig.cap="**Figure.** Growth and consumption curves for juvenile bass (J).
+```{r figure-grow-cons-J, eval=FALSE, fig.width=6, fig.height=6, fig.cap="**Figure.** Growth and consumption curves for juvenile bass (J).
 Different panels show the curves for varying values of harvest rate on adult bass (qE)."}
+
+library(data.table)
+
 # dFJ_dt_1state <- function(State0, pars, stateName=c("J0","F0"), parts=FALSE){

### 2018-12-11 paper/paper.Rmd
Starting on revisions
WIP

@@
 author:
-  - name: Ryan Batt
-    affiliation: rb
   - name: Carl Boettiger
@@
     footnote: Corresponding Author
+  - name: Ryan Batt
+    affiliation: rb
 address:
@@
   Ecosystems can experience sudden regime shifts due to a variety of mechanisms.
-Two of the ways a system can cross a tipping point include when a perturbation to the system state is large enough to push the system beyond the basin of attraction of one stable state and into that of another (state tipping), and alternately, when slow changes to some underlying parameter lead to a fold bifurcation that anihilates one of the stable states.
+Two of the ways a system can cross a tipping point include when a perturbation to the system state is large enough to push the system beyond the basin of attraction of one stable state and into that of another (state tipping), and alternately, when slow changes to some underlying parameter lead to a fold bifurcation that annihilates one of the stable states.
 The first mechanism does not generate the phenomenon of critical slowing down (CSD), whereas the latter does generate CSD, which has been postulated as a way to detect early warning signs ahead of a sudden shift.
@@
 Therefore, ecologists have often turned to studying the dynamics of nonlinear but simple mathematical models to better understand and manage ecosystems.
-There are essentially two different mechanisms by which a system exhibiting alternative stable states can experience a sudden transition from one state to another [@Scheffer2001]: state tipping (S-tipping) or a bifurcation (B-tipping).
+There are essentially two different mechanisms by which a system exhibiting alternative stable states can experience a sudden transition from one state to another [@Scheffer2001; @Ashwin2012]: state tipping (S-tipping) or a bifurcation (B-tipping).
 Figure \ref{fig1} illustrates these two  mechanisms.
@@
 S-tipping can be driven by direct exogenous forcing or purely stochastic events, such as extreme events that are common in the biological time series of ecosystems [@Batt2017].
+While real-world transitions will often be a combination of both processes, one mechanism will typically have a dominant role.
+For instance, near enough to the bifurcation, intrinsic noise will frequently drive the system across the tipping point before the actual bifurcation event is reached.
 Crucially, only one of these mechanisms, B-tipping, will exhibit the phenomenon of Critical Slowing Down (CSD), a pattern that has received great attention in both theoretical and empirical studies for its potential as an early warning signal of such a regime shift [e.g.
@@
 \includegraphics[width=.8\linewidth]{figures/scheffer-fig.png}
-\caption{S-tipping vs B-tipping. A critical transition in a system with
-alternative stable states can be driven either by a perturbation to the
-state, driving the system across the separatrix (top panel), or by a
-bifurcation, in which a stable point is annihilated (bottom panel). 
-Figure reproduced from @Scheffer2001 with permission.}
+\caption{S-tipping vs B-tipping.
+Both panels show the bifurcation diagram for a saddle-node or "fold" bifurcation.
+These plot the equilibrium points of a system on the vertical axis (e.g.
+adult bass density in our example) as a function of environmental conditions (parameters of the system).
+Solid lines indicate stable points, dotted lines indicate unstable points.
+The top panel illustrates Bifurcation Tipping: a system which begins in the desirable, "high" ecosystem state (top branch) experiences gradual degradation as the underlying environmental conditions change (increasing values along the $x$).
+At the F2 position, this stable node collides with the unstable saddle point (dotted line) and is annihilated, leaving the system to rapidly transition to the remaining undesirable stable state (Forward shift).
+To restore the system to the desirable state, it is not sufficient to merely restore the environmental conditions observed at F2.
+Instead, environmental conditions must improved all the way to F1, where another bifurcation occurs that creates a rapid backwards shift to the desirable state.
+This effect is known as hysteresis.
+The bottom panel introduces state tipping: the transition from the desirable state to the undesirable state is triggered without changing the environmental conditions.
+Instead, a large perturbation is introduced directly to the ecosystem state, sufficient to push the system from one basin-of-attraction to another.
+Such transitions are characterized by large, sudden shocks to the state variable, rather than slow, gradual change to the environmental conditions.
+These transitions do not exhibit hysteresis -- a sufficiently large perturbation could restore the system.
+Figure reproduced from Scheffer et al (2001) with permission.}\label{fig2}
 \end{figure}
@@
 Although the simplicity of ball-in-cup (potential well) based descriptions of regime shifts makes them alluring, this question is impossible to address properly without a more mathematically precise description of the system and the manipulation involved.
-Surprisingly, despite the central importance the piscivore transitions have played in our understanding of ecosystem regime shifts, these details have never been clearly presented in prior publications.
+Surprisingly, despite the central importance the piscivore transitions have played in our understanding of ecosystem regime shifts, these details have never been clearly presented in prior publications of this classic lake ecosystem experiment [@Carpenter2008; @Carpenter2011].
 Here, we provide a direct analysis of the underlying model in the piscivory alternative state system to demonstrate why these studies are indeed an example of B-tipping.
@@
 
-Decreases in adult bass abundance result from mortality ($m_A A$) and harvest ($q$), and increases result from the maturation of the fraction of $J$ that survive ($s$) mortality (Eq.
+Decreases in adult bass abundance result from mortality ($m_A A$) and harvest effort ($q$), and increases result from the maturation of the fraction of $J$ that survive ($s$) mortality (Eq.
 1).
+Stocking the lake corresponds to $q < 0$.
+This continuous-time formulation for harvest or stocking effort can be justified when carried out throughout the year at a fixed rate, as in the experiments considered here, rather than in large shocks.
 Planktivorous fish diffuse ($D_F$) between a refuge of size $F_o$ and a foraging arena, and the predation rate of $A$ on $F$ is controlled by $c_{FA}$ (Eq.
@@
 At each time step a small amount of additive, normally distributed noise was added to Eqs.
-1-3.
-Values for state variables less than 0.001 were set to 0.001 to prevent ecologically impossible values resulting from our numerical approximation of continuous dynamics.
-R code for these simulations is included in supplementary materials (<https://github.com/cboettig/bs-tipping).
+1-3, truncated to non-zero values.
+R code for these simulations is included in supplementary materials (<https://github.com/cboettig/bs-tipping>).
 
@@
 \centering
-\includegraphics[width=.8\linewidth]{../pkgBuild/manuscript_report/figure-5dsimulation-1.png}
+\includegraphics[width=.8\linewidth]{../analysis/manuscript_report/figure-5dsimulation-1.png}
 \caption{Time series of the 5 state variables, with \(q\) (harvest rate
@@
 These same dynamics can be seen in the direct numerical simulations of the full 5-dimensional model.
-At a $q$ of 0, the fish in the simulation underwent a critical transition (Figure \ref{fig4}).
+At a $q$ of 0, the fish in the simulation underwent a critical transition (Figure \ref{fig3}).
 This transition occurred during time step 510, although an abrupt change in state was not observed until step 850.
@@
 \begin{figure}
-\includegraphics[width=\linewidth]{../pkgBuild/manuscript_report/figure-Amotion-ballincup-1.png}
-\caption{Top row: motion of A for different values of $q$. Bottom row: the negative integral of the top row, depecting in the so-called ball-in-cup (or potential well) diagrams.} \label{fig4}
+\includegraphics[width=\linewidth]{../analysis/manuscript_report/figure-Amotion-ballincup-1.png}
+\caption{Top row: motion of A for different values of $q$.
+Bottom row: the negative integral of the top row, depicting in the so-called ball-in-cup (or potential well) diagrams.
+Note that the peaks and troughs (local maxima and minima) of the potential wells, bottom row, correspond to the zeros of the rate equations (top row).
+Troughs indicate stable nodes, peaks indicate unstable saddle points.
+Note that under sufficient stocking of bass, $q=-0.5$, the zero point is an unstable equilibrium, indicated by the peak in the potential well (bottom right).} \label{fig4}
 \end{figure}
@@
 \centering
-\includegraphics[width=.8\linewidth]{../pkgBuild/manuscript_report/figure-bifurcationDiagram-1.png}
+\includegraphics[width=.8\linewidth]{../analysis/manuscript_report/figure-bifurcationDiagram-1.png}
 \caption{Bifurcation diagram derived from Eq 9. Solid lines are stable

### 2018-12-11 paper/paper.Rmd
WIP

@@
 Central to resolving this question is untangling a second issue that frequently arises in the discussion of alternative stable state dynamics in ecological systems: the issue of dimension.
+Ecosystems are notoriously complex, and this lake ecosystem is no exception.
+Here, we analyze a previously published model of a trophic cascade to determine whether manipulating the top predator constitutes B-tipping or S-tipping.
+Common versions of this model include 5 state variables, ranging from top predators to primary producers (Figure \ref{fig2}).
+In sharp contast to this complexity, both intutition and theory of critical transitions have been centered on one-dimensional intpretations.
 The S-tipping and B-tipping transitions illustrated in Figure \ref{fig1} are fundamentally rooted in a one-dimensional analysis: the state consists of a single continuously valued dynamical variable (i.e.
 population density of adult bass).
-Ecosystems are notoriously more complex, and this lake ecosystem is no exception.
-Here, we analyze a previously published model of a trophic cascade to determine whether manipulating the top predator constitutes B-tipping or S-tipping.
-Common versions of this model include 5 state variables, ranging from top predators to primary producers (Figure \ref{fig2}).
-We perform dimension reduction on the full 5D model.
+Similarly, Ball-in-cup style depictions of alternative stable states and sudden transitions present only a one-dimensional dipiction of a system state.
+While it may be tempting to dismiss such simple depictions as out-of-hand as mere cartoons, it is in fact possible to map the full five-dimensional model explicitly into a one-dimensional approximation under a few simple assumptions.
+Here, we present the full dimension reduction on the full 5D model analytically, and compare this approximation to numerical simulations of the original system.
 This exercise facilitates the interpretation of the lake experiment in the context of predicting regime shifts by using a 1D model to understand a high dimensional system.
@@
 It is this aspect that first gives the appearance of state tipping dynamics, since it suggests the state is being somehow "directly manipulated".
-Yet from a mathematical standpoint, it is irrelevant if the control parameter corresponds to some environmental or policy variable (fishing regulations, fish migration into the lake), which in turn increases or decreases net mortality proportionally, or if the manipulation directly adds or removes adult bass -- the feature that distinguishes this manipulation from a S-tipping transition comes down to the timescales involved.
+Yet from a mathematical standpoint, it is irrelevant if the control parameter corresponds to some environmental or policy variable (fishing regulations, fish migration into the lake), which in turn increases or decreases net mortality proportionally, or if the manipulation directly adds or removes adult bass -- the feature that distinguishes this manipulation from a S-tipping transition comes down to the scales involved.
+This can be seen most clearly though an analogy of considering the same model applied to a larger scale, such as marine fisheries.
+In fisheries models, one would routinely examine the dynamics of the fish population biomass $B$ with respect to different possibly parameters for harvests or fishing effort.
+Even though harvest may be measured in the same units as the total biomass, it is natural to think of the two quantites very differently.
+A manager can adjust the harvest effort by changing quotas, limiting permits for fishing boats etc.
 Slowly increasing in the amount of bass stocked in the lake (i.e.

### 2018-12-14 paper/paper.Rmd
completed revisions

@@
 Common versions of this model include 5 state variables, ranging from top predators to primary producers (Figure \ref{fig2}).
-In sharp contast to this complexity, both intutition and theory of critical transitions have been centered on one-dimensional intpretations.
+In sharp contrast to this complexity, both intuition and theory of critical transitions have been centered on one-dimensional interpretations.
 The S-tipping and B-tipping transitions illustrated in Figure \ref{fig1} are fundamentally rooted in a one-dimensional analysis: the state consists of a single continuously valued dynamical variable (i.e.
 population density of adult bass).
-Similarly, Ball-in-cup style depictions of alternative stable states and sudden transitions present only a one-dimensional dipiction of a system state.
+Similarly, Ball-in-cup style depictions of alternative stable states and sudden transitions present only a one-dimensional depiction of a system state.
 While it may be tempting to dismiss such simple depictions as out-of-hand as mere cartoons, it is in fact possible to map the full five-dimensional model explicitly into a one-dimensional approximation under a few simple assumptions.
@@
 In fisheries models, one would routinely examine the dynamics of the fish population biomass $B$ with respect to different possibly parameters for harvests or fishing effort.
-Even though harvest may be measured in the same units as the total biomass, it is natural to think of the two quantites very differently.
+Even though harvest may be measured in the same units as the total biomass, it is natural to think of the two quantities very differently.
 A manager can adjust the harvest effort by changing quotas, limiting permits for fishing boats etc.
@@
 
-More generally, our analysis draws attention to issues of dimensionality that have been largely overlooked despite significant interest in questions of regime shifts and early warning signals [e.g.
-@Clements2018; @Scheffer2015; @Scheffer2012].
-The popularity of this topic has sometimes led to casually equating any sudden shift with fold bifurcations, overlooking any assumptions necessary to approximate a complex dynamical system with such a one-dimensional model.
+A second issue highlighted by this analysis is the contrast between the interpretation of the five-dimensional food web model and the classic one-dimensional description of tipping points shown in Fig \ref{fig1}.
+Such one-dimensional descriptions of a state are frequently presented in reviews or abstract discussion of regime shifts and early warnings with little or no discussion of how a complex system comes to be represented by a one-dimensional continuous variable [e.g.
+@Scheffer2001; @Scheffer2009; @Barnosky2012; @Scheffer2012; @Clements2018], and even literature focusing on specific systems such as the lake-ecosystem  considered here can be vague about the dimensional reduction [@Carpenter2008; @Carpenter2011].
+This can in turn create confusion as to whether such one-dimensional diagrams are meant to be interpreted as mathematically explicit approximations of higher-order dynamics, or merely as cartoons meant only to mimic the qualitative pattern but bearing no quantitative relationship with the actual process.
+The analysis presented here for the lake-ecosystem model demonstrates that the former does indeed hold, though only under the appropriate time-scale limits -- not only that of the "slow parameter" but also that of the other dynamic variables in the system: plankton, Juvenile Bass, Forage Fish, Fig \ref{fig2}.
+Our analysis also highlights that how the nonlinearity required to create a fold bifurcation arises directly from this dimension-reducing approximation.
+In the original five-dimensional model, Eq (1), the dynamics of the adult bass $A$ appear to be completely linear.
+Only after accounting for the feedback in the other dimensions do we see the nonlinearity required to create a fold bifurcation.
+Our numerical simulations provide an illustration of how closely the dynamics of the full five-dimensional system are captured by the one-dimensional approximation using a biologically plausible parameterization.
+Thus the reference to the dynamics of a one-dimensional system state, as well as the interpretation of the transition as B-tipping rather than S-tipping, appears to be well-justified in the lake-ecosystem case.
+
+This analysis does not vindicate the use of one-dimensional models more generally.
+The popularity and simplicity of the fold-bifurcation model has sometimes led to more casually equating any sudden shift with fold bifurcations, overlooking any assumptions necessary to approximate a complex dynamical system with such a one-dimensional model.
 Influential reviews on abrupt change in ecological systems [e.g.
-@Ratajczak2018] focus almost exclusively on descriptions of one-dimensional state-spaces without discussion of if, when or how complex systems such as the ecosystem food-web considered here can be appropriately described by a nonlinear equation in a single dimension.
-The analysis shown here illustrates this process on the basis of a fast-slow timescale approximation, which allows us to replace the faster dynamics at lower trophic levels with their equilibrium states.
-(Note also that the nonlinearity necessary to support alternate stable states only emerges through this approximation.)  Thus, for the five-dimensional model considered here, our analysis shows that the familiar one-dimensional characterization of a sudden shift being driven by a fold bifurcation is indeed appropriate, but this was not altogether evident a priori.
+@Scheffer2015; @Ratajczak2018] focus almost exclusively on descriptions of one-dimensional state-spaces without discussion of if, when or how complex systems such as the ecosystem food-web considered here can be appropriately described by a nonlinear equation in a single dimension.
+More careful modeling informed by system-specific knowledge is required before any abrupt transition can be characterized as a fold bifurcation [@Nature2013].
 The fact that CSD phenomena can occur in many different models has been suggested to mean that CSD is a generic or universal early warning indicator [@Scheffer2009] of a critical transition.
